# Introduction {#intro}



## What is R? {#what-r}

R was not designed to be a bona-fide programming language. 
It is an evolution of the S language, developped at Bell labs (later Lucent) as a wrapper for the endless collection of statistical libraries they wrote in Fortran.

As of 2011, half of R's libraries are [actually written in C](https://wrathematics.github.io/2011/08/27/how-much-of-r-is-written-in-r/). 



For more on the history of R see [AT&T's site](http://www.research.att.com/articles/featured_stories/2013_09/201309_SandR.html?fbid=Yxy4qyQzmMa), John Chamber's talk at [UserR! 2014](https://www.youtube.com/watch?v=_hcpuRB5nGs) or the Introduction to the excellent @venables2013modern.

## The R Ecosystem {#ecosystem}

A large part of R's success is due to the ease in which a user, or a firm, can augment it. 
This led to a large community of users, developpers, and protagonists. 
Some of the most important parts of R's ecosystem include:

- [CRAN](https://cran.r-project.org/): 
a repository for R pcakges, mirrored worldwide.

- [R-help](https://www.r-project.org/mail.html): 
an immensly active mailing list. 
Nowdays being replaced by StackExchange meta-site. 
Look for the R tags in the [StackOverflow](http://stackoverflow.com/) and [CrossValidated](http://stats.stackexchange.com/) sites.

- [TakViews](https://cran.r-project.org/web/views/): 
part of CRAN that collects packages per topic. 
- [Bioconductor](https://www.bioconductor.org/): 
A CRAN-like repository dedicated to the life sciences. 

- [Books](https://www.r-project.org/doc/bib/R-books.html): 
An insane amount of books written on the language. Some are free, some are not. 

- [The Israeli-R-user-group](https://groups.google.com/forum/#!forum/israel-r-user-group): 
just like the name suggests.

- Commercial R: being open source and lacking support may seem like a problem that would prehibit R from being adopted for commercial applications. 
This void is filled by several very succesfull commercial versions such as [Microsoft R](https://mran.microsoft.com/open/), with its accompanying CRAN equivalent called [MRAN](https://mran.microsoft.com/),  [Tibco's Spotfire](http://spotfire.tibco.com/discover-spotfire/what-does-spotfire-do/predictive-analytics/tibco-enterprise-runtime-for-r-terr), and [others](https://en.wikipedia.org/wiki/R_(programming_language)#Commercial_support_for_R).

- [RStudio](https://www.rstudio.com/products/rstudio/download-server/): since its easliest days R came equipped with a minimal text editor. 
It later recieved plugins for major integrated developpment environments (IDEs) such as Eclipse, WinEdit and even [VisualStudio](https://www.visualstudio.com/vs/rtvs/). 
None of these, however, had the impact of the RStudio IDE. 
Written completely in JavaScript, the RStudio IDE allows the seamless integration of cutting edge web-design technologies, remote access, and other killer features, making it todays most popular IDE for R.



## R Basics 

We now start with the basics of R. 
If you have any experience at all with R, you can probably skip this section.

First, make sure you work with the RStudio IDE.
Some useful pointers for this IDE include:
- Ctrl+return to run lines from editor.
- alt+shift+k for RStudio keyboard shortcuts.
- Ctrl+alt+j to navigate between sections
- tab for auto-completion
- Ctrl+1 to skip to editor. 
- Ctrl+2 to skip to console.
- Ctrl+8 to skip to the environment list.
- Code Folding:
  - alt+l collapse chunk.
  - alt+shift+l unfold chunk.
  - alt+o collapse all.
  - alt+shift+o unfold all.


### Simple calculator
R can be used as a simple calculator.
```{r}
10+5
70*81
2**4
2^4
log(10)       					
log(16, 2)    					
log(1000, 10)   				
```


### Probability calculator 
R can be used as a probability calculator. 
You probably wish you knew this when you did your Intro To Probability.

The binomial distribution function:
```{r}
dbinom(x=3, size=10, prob=0.5) 	# Compute P(X=3) for X~B(n=10, p=0.5) 
```
Notice that arguments do not need to be named explicitely
```{r}
dbinom(3, 10, 0.5)
```
The binomial cummulative distribution function (CDF):
```{r}
pbinom(q=3, size=10, prob=0.5) # Compute P(X<=3) for X~B(n=10, p=0.5)  	
```
The binomial quantil function:
```{r}
qbinom(p=0.1718, size=10, prob=0.5) # For X~B(n=10, p=0.5) returns k such that P(X<=k)=0.1718
```
Generate random variables:
```{r}
rbinom(n=1, size=10, prob=0.5) 	
rbinom(n=10, size=10, prob=0.5)
rbinom(n=100, size=10, prob=0.5)
```

R has many built-in distributions. 
Their names may change, but the prefixed do not:

- __d__ prefix for the distribution function.
- __p__ prefix for the CDF.
- __q__ prefix for the quantile function (i.e., the inverse CDF).
- __r__ prefix to generate sandom samples. 



## Getting help
One of the most important parts of working with a language, is to know where to find help. 
R has several in-line facilities, besides the various help resources in the R [ecosystem](#ecosystem).


Get help for a particular function.
```{r, eval=FALSE}
?dbinom 
help(dbinom)
```

Search local help files for a particular string.
```{r, eval=FALSE}
??binomial
help.search('dbinom') 
```

Load a menu with several important manuals:
```{r, eval=FALSE}
help.start() 
```


## Variable asignment:
Assignments into a variable named "x":
```{r}
x = rbinom(n=1000, size=10, prob=0.5) # Works. Bad style.
x <- rbinom(n=1000, size=10, prob=0.5) # Asignments into a variable named "x"
```
More on style: http://adv-r.had.co.nz/Style.html


Print contents:
```{r}
x
print(x)  
(x <- rbinom(n=1000, size=10, prob=0.5))  # Assign and print.
```


Operate on the object
```{r}
mean(x)  
var(x)  
hist(x)  
rm(x) # remove variable
```


For more information on distributions see http://cran.r-project.org/web/views/Distributions.html
