# R Basics {#basics}

We now start with the basics of R. 
If you have any experience at all with R, you can probably skip this section.

First, make sure you work with the RStudio IDE.
Some useful pointers for this IDE include:
- Ctrl+return to run lines from editor.
- alt+shift+k for RStudio keyboard shortcuts.
- Ctrl+alt+j to navigate between sections
- tab for auto-completion
- Ctrl+1 to skip to editor. 
- Ctrl+2 to skip to console.
- Ctrl+8 to skip to the environment list.
- Code Folding:
  - alt+l collapse chunk.
  - alt+shift+l unfold chunk.
  - alt+o collapse all.
  - alt+shift+o unfold all.


## Simple calculator
R can be used as a simple calculator.
```{r}
10+5
70*81
2**4
2^4
log(10)       					
log(16, 2)    					
log(1000, 10)   				
```


## Probability calculator 
R can be used as a probability calculator. 
You probably wish you knew this when you did your Intro To Probability.

The binomial distribution function:
```{r}
dbinom(x=3, size=10, prob=0.5) 	# Compute P(X=3) for X~B(n=10, p=0.5) 
```
Notice that arguments do not need to be named explicitly
```{r}
dbinom(3, 10, 0.5)
```
The binomial cumulative distribution function (CDF):
```{r}
pbinom(q=3, size=10, prob=0.5) # Compute P(X<=3) for X~B(n=10, p=0.5)  	
```
The binomial quantile function:
```{r}
qbinom(p=0.1718, size=10, prob=0.5) # For X~B(n=10, p=0.5) returns k such that P(X<=k)=0.1718
```
Generate random variables:
```{r}
rbinom(n=10, size=10, prob=0.5)
```

R has many built-in distributions. 
Their names may change, but the prefixed do not:

- __d__ prefix for the distribution function.
- __p__ prefix for the CDF.
- __q__ prefix for the quantile function (i.e., the inverse CDF).
- __r__ prefix to generate random samples. 



## Getting Help
One of the most important parts of working with a language, is to know where to find help. 
R has several in-line facilities, besides the various help resources in the R [ecosystem](#ecosystem).


Get help for a particular function.
```{r, eval=FALSE}
?dbinom 
help(dbinom)
```
If you don't know the name of the function you are looking for, search local help files for a particular string:
```{r, eval=FALSE}
??binomial
help.search('dbinom') 
```
Or load a menu where you can navigate local help in a web-based fashion:
```{r, eval=FALSE}
help.start() 
```


## Variable Asignment
Assignment of some output into an object named "x":
```{r}
x = rbinom(n=10, size=10, prob=0.5) # Works. Bad style.
x <- rbinom(n=10, size=10, prob=0.5) 
```
If you are familiar with other programming languages you may prefer the '=' assignment rather than the '<-' assignment. 
We recommend you make the effort to change your preferences. 
This is because thinking with '<-' helps to read your code, distinguishes between assignments and function arguments: `function(argument=value)`, and understand things like `<<-` and `->`. 


```{remark}
__Style__:
We do not discuss style guidelines in this text, but merely remind the reader that good style is extremely important. When you write code, think of other readers, but also think of future self.
See [Hadley's style guide](http://adv-r.had.co.nz/Style.html) for more. 
```

To print the contents of an object just type its name
```{r}
x
```
which is an implicit call to 
```{r}
print(x)  
```
Alternatively, you can assign and print simultaneously
```{r}
(x <- rbinom(n=10, size=10, prob=0.5))  # Assign and print.
```

Operate on the object
```{r}
mean(x)  # compute mean
var(x)  # compute variance
hist(x) # plot histogram
```

R saves every object you create in RAM^[S and S-Plus used to save objects on disk. Working from RAM has advantages and disadvantages. More on this in Chapter \@ref(memory).]. 
The collection of all such objects is the __workspace__ which you can inspect with 
```{r}
ls()
```
or with Ctrl+8 in RStudio. 

If you lost your object, you can use `ls` with a text patter to search for
```{r}
ls(pattern='x')
```



To remove objects from the workspace:
```{r}
rm(x) # remove variable
ls() # verify
```
You may think that if an object is removed then its memory is freed. 
This is almost true, and depends on a negotiation mechanism between R and the operating system. 
R's memory management is discussed in Chapter \@ref(memory). 


## Piping 
Because R originates in Unix and Linux environments, it inherits much of its flavor.
[Piping](http://ryanstutorials.net/linuxtutorial/piping.php) is an idea take from the Linux shell which allows to use the output of one expression as the input to another.
Piping thus makes code easier to read and write. 

```{remark}
Volleyball fans may be confused with the idea of spiking a ball from the 3-meter line, also called [piping](https://www.youtube.com/watch?v=GWW15Nr1lQM). 
So:
(a) These are very different things.
(b) If you can pipe, [ASA-BGU](http://in.bgu.ac.il/sport/Pages/asa.aspx) is looking for you!
```

Prerequisites:
```{r}
library(magrittr)
x <- rbinom(n=1000, size=10, prob=0.5)
```

Examples
```{r, eval=FALSE}
x %>% var() # Instead of var(x)
x %>% hist()  # Instead of hist(x)
x %>% mean() %>% round(2) %>% add(10) 
```


The next example^[Taken from  http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html] demonstrates the benefits of piping.
The next two chunks of code do the same thing.
Try parsing them in your mind:
```{r, eval=FALSE}
# Functional (onion) style
car_data <- 
  transform(aggregate(. ~ cyl, 
                      data = subset(mtcars, hp > 100), 
                      FUN = function(x) round(mean(x, 2))), 
            kpl = mpg*0.4251)
```

```{r, eval=FALSE}
# Piping (magrittr) style
car_data <- 
  mtcars %>%
  subset(hp > 100) %>%
  aggregate(. ~ cyl, data = ., FUN = . %>% mean %>% round(2)) %>%
  transform(kpl = mpg %>% multiply_by(0.4251)) %>%
  print
```



## Vector creation and manipulation 
The most basic building block in R is the __vector__.
We will now see how to create them, and access their elements (i.e. subsetting).
Here are three ways to create the same arbitrary vector:
```{r, eval=FALSE}
c(10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21) # manually
10:21 # the `:` operator 							
seq(from=10, to=21, by=1) # the seq() function
```
Lets assign it to the object named "x":
```{r}
x <- c(10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21) 	
```
In the case you made a computation you do not want to repeat, you can assign AFTER the computation is finished, since everything is saved by the `.Last.value' variable. 
```{r, results='hold'}
c(1,2,3)
y<- .Last.value 
y
```

```{remark}
In line with the linux look and feel, variables starting with a dot (.) are saved but are hidden.
To show them see `?ls`.
```


Operations usually work element-wise:
```{r}
x+2
x*2    
x^2    
sqrt(x)  
log(x)   
```


## Search paths and packages
R can be easily extended with packages, which are merely a set of functions and other objects, which can be loaded or unloaded at will.
Let's look at the function `sum`. 
We can see its contents by calling it without arguments:
```{r}
print(read.csv)
```
Never mind what the function does. 
Note the `environment: namespace:utils` line at the end. 
It tells us that this function is part of the __utils__ package.
We did not need to know this because it is loaded by default.
Here are the packages that are currently loaded:
```{r}
search()
```
Other packages can be loaded via the `library` function, or downloaded from the internet using the `install.packages` function before loading with `library`. 
R's package import mechanism is quite powerful, and is one of the reasons for R's success. 



## Simple plotting 
R has many plotting facilities. 
We start with the simplest facilities, namely, the `plot` function from the __graphics__ package, which is loaded by default. 
```{r}
x<- 1:100; y<- 3+sin(x) # Create arbitrary data
plot(x = x, y = y) # x,y syntax  						
```
Given an x argument and a y argument, `plot` tries to present a scatter plot.
We call this the "x,y" syntax.
R has another, unique, syntax to state functional relations.
We call it the "tilde" syntax, which originates in works of @wilkinson1973symbolic. 
```{r}
plot(y ~ x) # y~x syntax 
```

The syntax `y~x` is read as "y is a function of x".
We will prefer the `y~x` syntax over the `x,y` syntax since it is easier to read, and will be very useful when we discuss more complicated models. 




Here are some arguments that control the plot's appearance:
```{r}
plot(y~x, type='l', main='Plotting a connected line') # main title
plot(y~x, type='h', main='Sticks plot', xlab='Insert x axis label', ylab='Insert y axis label') # axes labels
plot(y~x, pch=5) # Point type with pcf
plot(y~x, pch=10, type='p', col='blue', cex=4) # More point parameters
abline(3, 0.002) # add linear line with slope b and intercept a
```

For more plotting options run these
```{r, eval=FALSE}
example(plot)
example(points)
?plot
help(package='graphics')
```

When your plotting gets serious, go to Chapter \@ref(plotting).


## Object types
We already saw that the basic building block of R objects is the vector.
Vectors can be of the following types:

- __character__ Where each element is a string. 
- __numeric__ Where each element is a "real" number in double precision floating point. 
- __integer__ Where each element is an integer. 
- __logical__ Where each element is either TRUE, FALSE, or NA^[R uses a [three valued logic](https://en.wikipedia.org/wiki/Three-valued_logic) where a missing value (NA) is neither TRUE, nor FALSE.]
- __complex__ Where each element is a complex number. 
- __list__ Where each element is an arbitrary R object.
- __factor__ Factors are not actually vector objects, but they feel like such. 
They actually used to encode a finite set of values. 
This will be very useful when fitting linear model, but may be confusing if you think you are dealing with a character vector when in fact you are dealing with a factor. 
Be alert!

Vectors can be combined into larger objects. 
A `matrix` can be thought of as the binding of several vectors of the same type. 
If vectors of different types (but same length) are binded, we get a `data frame` which is the most fundamental object in R for data analysis. 



## Data Frames 
Creating a simple data frame:
```{r, results='hold'}
x<- 1:10; y<- 3 + sin(x) 
frame1 <- data.frame(x=x, sin=y)	
```

Lets inspect our data frame:
```{r}
head(frame1)
```

Now using the RStudio Excel-like viewer:
```{r, eval=FALSE}
frame1 %>% View() 
```
We highly advise against editing the data this way since there will be no documentation of the changes you made.

Verifying this is a data frame:
```{r}
class(frame1) # the object is of type data.frame
```

Check the dimension of the data
```{r}
dim(frame1)  							
```

Note that checking the dimension of a vector is different than checking the dimension of a data frame.
```{r}
length(x)
```
A frame is a vector of column vectors, so its length is merely the number of columns.
```{r}
length(frame1) 
```


## Exctraction
R provides many ways to subset and extract elements from vectors and other objects. 
The basics are fairly simple, but not paying attention to the "personality" of each extraction mechanism may cause you a lot of headache. 

For starters, extraction is done with the `[` operator. 
The operator can take vectors of all types. 

Extracting element with by integer index:
```{r}
frame1[1, 2]  # exctract the element in the 1st row and 2nd column.
```

Extract __column__ by index:
```{r}
frame1[1, ]      						
```

Extract column by name:
```{r}
frame1[, 'sin']
```

What did we just extract?


```{r}
dim(frame1[, 'sin'])  # extracts a column vector
dim(frame1['sin'])  # extracts a data frame
dim(frame1[,1:2])  # extracts a data frame
dim(frame1[2])  # extracts a data frame
dim(frame1[2, ])  # extract a data frame
dim(frame1$sin)  # extracts a column vector
```


The `subset()` function does the same
```{r, eval=FALSE}
subset(frame1, select=sin) 
subset(frame1, select=2)
subset(frame1, select= c(2,0))
```


If you are unsatisfied with the output of the `[` mechanism, you can use `[[`, which gets the content of a vector, while stripping the attributes.
```{r}
a <- frame1[1] # [ extraction
b <- frame1[[1]] # [[ extraction
a==b # objects are element-wise identical 
class(a)==class(b)
```

The different types of output causes different behaviors
```{r}
a[1]
b[1]
```

If you want to learn more about subsetting see [Hadley's guide](http://adv-r.had.co.nz/Subsetting.html), and our Chapter \@ref(hadley)






## Data Import and Export 

For any practical purpose, you will not be generating your data manually. 
R comes with many importing and exporting mechanism which we now present.
If, however, you do a lot of data "munging", make sure to see Hadley-verse Chapter \@ref(hadley).
If you work with MASSIVE data sets, read about the __data.table__ package. 
For a complete review see the [R manual](http://cran.r-project.org/doc/manuals/R-data.html).


### Import from WEB 
The `read.table` function is the main importing workhorse.
It can import directly from the web.
```{r}
URL <- 'http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/bone.data'
tirgul1 <- read.table(URL)
```

Always look at the imported result!
```{r}
head(tirgul1)
```

Ohh dear. The header row was not recognized. Fix with `header=TRUE`:
```{r}
tirgul1 <- read.table(URL, header = TRUE) 
head(tirgul1)
```



### Export as CSV 
Let's write a simple file so that we have something to import
```{r}
head(airquality) #  examine the data to export
temp.file.name <- tempfile() # get some arbitrary file name
write.csv(x = airquality, file = temp.file.name) # export
```
Now let's import the exported file. Being a .csv file, I can use `read.csv` instead of `read.table`. 
```{r}
my.data<- read.csv(file=temp.file.name) # import
head(my.data) # verify import
```


```{remark}
Windows users may need to use "\\" instead of "/".
```



### Reading From Text Files 

Some general notes on importing text files via the `read.table` function.
But first, we need to know what is the active directory.
Here is how to get and set R's active directory:
```{r, eval=FALSE}
getwd() #What is the working directory?
setwd() #Setting the working directory in Linux
```

We can now call the `read.table` function to import text files. 
If you care about your sanity, see `?read.table` before starting imports.
Some notable properties of the function:

- `read.table` will try to guess column separators (tab, comma, etc.)
- `read.table` will try to guess if a header row is present.
- `read.table` will convert character vectors to factors unless told not to.
- The output of `read.table` needs to be explicitly assigned to an object for it to be saved. 



### Writing Data to Text Files

The function `write.table` is the exporting counterpart of `read.table`.


### .XLS(X) files 
Strongly recommended to convert to .csv in Excel, and then import as csv. 
If you still insist see [here](http://cran.r-project.org/doc/manuals/R-data.html#Reading-Excel-spreadsheets).


### Massive files 

The above importing and exporting mechanism were not designed for massive files. 
See the section on Sparse Representation (\@ref(sparse)) and Out-of-Ram Algorithms (\@ref(memory)) for more on working with massive data files.  


### Databases 
R can does not need to read from text files; it can read directly from a data base.
This is very useful since it allows the filtering, selecting and joining operations to rely on the database's optimized algorithms.
See [here](https://rforanalytics.wordpress.com/useful-links-for-r/odbc-databases-for-r/).




## Functions
One of the most basic building blocks of programming is the ability of writing your own functions.
A function in R, like everything else, is a an object accessible using its name.
We first define a simple function that sums its two arguments
```{r}
my.sum <- function(x,y) {
  x+y
}
my.sum(10,2)
```

From this example you may notice that:

- The function `function` tells R to construct a function object.
- The arguments of the `function`, i.e. `(x,y)`, need to be named but we are not required to specify their type. 
- A typical R function does not change objects^[This is a classical _functional programming_ paradigm. If you are used to _object oriented_ programming, you may want to read about [references classes](http://adv-r.had.co.nz/R5.html) which may be required if you are planning to compute with very complicated objects.] but rather creates new ones. 
To save the output pf `my.sum` we will need to assign it using the `<-` operator.
- The function will output its last evaluated expression. 



## Looping 
The real power of scripting is when repeated operations are done by iteration.
R supports the usual `for`, `while`, and `repated` loops. 
Here is an embarrassingly simple example
```{r looping}
for (i in 1:5){
    print(i)
    }
```


## Recursion 
The R compiler is really not designed for recursion, and you will rarely need to do so.  
See the RCpp Chapter \@ref(rcpp) for linking C code, which is better suited for recursion. 
If you really insist to write recursions in R, make sure to use the `Recall` function, as this 
Fibonacci series example demonstrates. 
```{r recusrion}
fib<-function(n) {
    if (n < 2) fn<-1 
    else fn<-Recall(n - 1) + Recall(n - 2) 
    return(fn)
} 
fib(5)
```



## Bibliographic Notes

There are endlessly many introductory texts on R.
I personally recommend the official introduction @venables2004introduction, or anything else Bill Venables writes. 
For advanced R programming see @wickham2014advanced, or anything else Hadley Wickham writes. 
