# Linear Mixed Models {#lme}

```{example, label='fixed-effects'}
Consider the problem of testing for a change in the distribution of the bottle caps produced.
Bottle caps are produced by several machines. 
We could standardize by removing each machine's average. 
This first practice implies the within-machine variability is the only source of variability. 
Alternatively, we could ignore the machine of origin. 
This second practice implies there are two sources of variability: the within-machine variability, and the between-machine variability.
The former practice is known as a _fixed effects_  model.
The latter as a _random effects_ model.
```


```{example, label='random-effects'}
Consider a crossover^[If you are unfamiliar with design of experiments, have a look at Chapter 6 of my Quality Engineering [class notes](https://github.com/johnros/qualityEngineering/blob/master/Class_notes/notes.pdf).] experimenal design where each subject is given 2 types of diets, and his/hers health condition is recorded.
We could standardize by removing each subject's average, before comparing the diets (think of a paired t-test).
This first practice implies the within-subject variability is the only source of variability.
Alternatively, we could ignore the subject of origin. 
When doing so, we need to recall that observations from the same subject will be correlated.
This second practice implies there are two sources of variability: the within-subject variability and the betwee-subject variability.
```

The unifying theme of the above two examples, is that the variability we want to infer against has several sources. 
This is typical in mixed models, which are so popular, that they have earned many names:

- __Mixed Effects__: 
Because we may have both _fixed effects_ we want to estimate and remove, and _random effects_ which contribute to the variability. 
- __Variance Components__: 
Because as the examples show, variance has more than a single source (like in the Linear Models of Chapter \@ref(lm)).
- __Hirarchial Models__: 
Because as Example \@ref(ex:random-effects) demonstrates, we can think of the sampling as hierarchical-- first sample a subject, and then sample its response. 
- __Repeated Measures__: 
Because we many have several measurements from each unit, like in \@ref(ex:random-effects).
- __Longitudinal Data__: 
Because we follow units over time, like in Example \@ref(ex:random-effects).
- __Panel Data__:
Is the term typically used in econometric for such longitudinal data. 


We now emphasize: 

1. Mixed effect models are a way to infer against the right level of variability.
Using a naive linear model (which assumes a single source of variability) instead of a mixed effects model, probably means your inference is overly anti-conservative, i.e., error rates are higher than you think. 
1. A mixed effect models, as we will later see, is typically specified via its fixed and random effects. 
It is possible, however, to specify a mixed effects model by putting all the fixed effects into a linear model, and putting all the random effects into the covariance between $\varepsilon$.
For more on this view, see Chapter 8 in (the excellent) @weiss2005modeling.
1. Like in previous chapters, by "model" we refer to the assumed generative distribution, i.e., the sampling distribution. 
1. If you are using the model merely for predictions, and not for inference on the fixed effects or variance components, then stating the generative distribution may be be useful, but not necessarily. 
See the Supervised Learning Chapter \@ref(supervised) for more on prediction problems.


## Problem Setup

\begin{align}
  y|x,u = x'\beta + z'u + \varepsilon
  (\#eq:mixed-model)  
\end{align}
where $x$ are the factor with fixed effects, $\beta$ which we may want to study.
The factors $z$, with effects $u$, are the random effects which contribute to variability. 
Put differently, we state $y|x,u$ merely as a convenient way to do inference on $y|x$, instead of directly specifying $Var[y|x]$ as a function of $u$. 

Given a sample of $n$ observations $(y_i,x_i,z_i)$ from model \@ref(eq:mixed-model), we will want to estimate $(\beta,u)$.
Under some assumption on the distribution of $\varepsilon$ and $z$, we can use _maximum likelihood_ (ML). 
In the context of mixed-models, however, ML is typically replaced with _restricted maximum likelihood_ (ReML), because it returns unbiased estimates of $Var[y|x]$ and ML does not.




## Mixed Models with R

We will fit mixed models with the `lmer` function from the __lme4__ package. 
We start with a small simulation demonstrating the importance of acknowledging your sources of variability.

```{r lme vs lm}
n.groups <- 10
n.repeats <- 2
groups <- gl(n = n.groups, k = n.repeats)
n <- length(groups)
z0 <- rnorm(10,0,10)
z <- z0[as.numeric(groups)] # create the random effect vector.
epsilon <- rnorm(n,0,1) # create the measurement error vector.
beta0 <- 2 # create the global mean
y <- beta0 + z  + epsilon # generate synthetic sample
lm.5 <- lm(y~z)  # fit a linear model
library(lme4)
lme.5 <- lmer(y~1|z) # fit a mixed-model
```

The summary of the linear model

```{r, label='lm5'}
summary.lm.5 <- summary(lm.5)
summary.lm.5
```

The summary of the mixed-model

```{r, label='lme5'}
summary.lme.5 <- summary(lme.5)
summary.lme.5
```
Look at the standard error of the global mean, i.e., the intercept:
for `lm` it is `r summary.lm.5$coefficients[1,2]`, and for `lme` it is `summary.lme.5$coefficients[1,2]`.
Why this difference? 
Because `lm` discounts the group effect, while it should treat it as another source of variability.
Clearly, inference using `lm` is overly optimistic.

### A Single Random Effect

We will use the `Dyestuff` data from the package, which encodes the yield, in grams, of a coloring solution (dyestuff), produced in 6 batches using 5 different preparations.

```{r}
data(Dyestuff, package='lme4')
attach(Dyestuff)
head(Dyestuff)
```

And visually

```{r}
plot(Yield~Batch)
```

If we want to do inference on the mean yield, we need to account for the two sources of variability: the batch effect, and the measurement error. 
We thus fit a mixed model, with an intercept and random batch effect (which means this is it not a bona-fide mixed-model, but rather, a simple random-effect model).

```{r random intercept}
lme.1<- lmer( Yield ~ 1  | Batch  , Dyestuff )
summary(lme.1)
```

Things to note:

- As usual, `summary` is content aware and has a different behavior for `lme` class objects.
- The syntax `Yield ~ 1  | Batch` tells R to fit a model with a global intercept (`1`) and a random Batch effect (`|Batch`). More on that later. 
- The output distinguishes between random effects, a source of variability, and fixed effect, white's coefficients we want to study.
- Were we not interested in the variance components, an (almost) equivalent `lm` formulation is `lm(Yield ~ Batch)`.

Some utility functions let us query the `lme` object. 
The function `coef` will work, but will return a cumbersome output. Better use `fixef` to extract the fixed effects, and `ranef` to extract the random effects.
The model matrix (of the fixed effects alone), can be extracted with `model.matrix`, and predictions made with `predict`.
Note, however, that predictions with mixed-effect models are (i) a delicate matter, and (ii) better treated as prediction problems as in the Supervised Learning Chapter \@ref(supervised).


### Several Random Effects

Let's make things more interesting.
In the `Penicillin` data, we measured the diameter of spread of an organism, along the plate used (a to x), and penicillin type (A to F). 

```{r}
detach(Dyestuff)
head(Penicillin)
```

One sample per combination:

```{r}
attach(Penicillin)
table(sample, plate)
```

And visually:

```{r}
lattice::dotplot(reorder(plate, diameter) ~ diameter,data=Penicillin,
              groups = sample,
              ylab = "Plate", xlab = "Diameter of growth inhibition zone (mm)",
              type = c("p", "a"), auto.key = list(columns = 6, lines = TRUE))
```

Let's fit a mixed-effects model with a random plate effect, and a random sample effect:

```{r}
lme.2 <- lmer ( diameter ~  1+ (1| plate ) + (1| sample ) , Penicillin )
fixef(lme.2) # Fixed effects
ranef(lme.2) # Random effects
```

Things to note:

- The syntax `1+ (1| plate ) + (1| sample )` fits a global intercept (mean), a random plate effect, and a random sample effect.
- Were we not interested in the variance components, an (almost) equivalent `lm` formulation is `lm(diameter ~ plate + sample)`.



Since we have two random effect, we may compute the variability of the global mean (the only fixed effect) as we did before. 
Perhaps more interestingly, we can compute the variability in the response, for a particular plate or sample type.

```{r}
random.effect.lme2 <- ranef(lme.2, condVar = TRUE) 
qrr2 <- lattice::dotplot(random.effect.lme2, strip = FALSE)
```

Things to note:

- The `condVar` argument of the `ranef` function tells R to compute the variability in response conditional on each random effect at a time. 
- The `dotplot` function, from the __lattice__ package, is only there for the fancy plotting.

Variability in response for each plate, over various sample types:

```{r}
print(qrr2[[1]]) 
```

Variability in response for each sample type, over the various plates:

```{r}
print(qrr2[[2]])  
```

### A Full Mixed-Model

In the `sleepstudy` data, we recorded the reaction times to a series of tests (`Reaction`), after various subject (`Subject`) underwent various amounts of sleep deprivation (`Day`).

```{r}
data(sleepstudy)
lattice::xyplot(Reaction ~ Days | Subject, sleepstudy, aspect = "xy",
             layout = c(9,2), type = c("g", "p", "r"),
             index.cond = function(x,y) coef(lm(y ~ x))[1],
             xlab = "Days of sleep deprivation",
             ylab = "Average reaction time (ms)")
```

We now want to estimate the (fixed) effect of the days of deprivation, while allowing each subject to have his/hers own effect.
The fixed effect can thus be thought of as the average slope over subjects.

```{r random slope}
lme.3 <- lmer ( Reaction ~ Days + ( Days | Subject ) , data= sleepstudy )
```

Things to note:

- We used the `Days|Subect` syntax to tell R we want to fit the model `~Days` within each subject.
- Were we fitting the model for purposes of prediction only, an (almost) equivalent `lm`formulation is `lm(Reaction~Days*Subject)`.


The fixed (i.e. average) day effect is:

```{r}
fixef(lme.3)
```

The variability in the average response (intercept) and day effect is

```{r}
ranef(lme.3)
```

Did we really need the whole `lme` machinery to fit a within-subject linear regression and then average over subjects?
The answer is yes.
The assumptions on the distribution of random effect, namely, that they are normally distributed, allows us to pool information from one subject to another. In the words of John Tukey: "we borrow strength over subjects".
Is this a good thing? If the normality assumption is true, it certainly is.

To demonstrate the "strength borrowing", here is a comparison of the subject-wise intercepts of the mixed-model, versus a subject-wise linear model. They are not the same.

```{r, echo=FALSE}
library(lattice)
df <- coef(lmList(Reaction ~ Days | Subject, sleepstudy))
fclow <- subset(df, `(Intercept)` < 251)
fchigh <- subset(df, `(Intercept)` > 251)
cc1 <- as.data.frame(coef(lme.3)$Subject)
names(cc1) <- c("A", "B")
df <- cbind(df, cc1)
ff <- fixef(lme.3)
with(df,
     print(xyplot(`(Intercept)` ~ Days, aspect = 1,
                  x1 = B, y1 = A,
                  panel = function(x, y, x1, y1, subscripts, ...) {
                    panel.grid(h = -1, v = -1)
                    x1 <- x1[subscripts]
                    y1 <- y1[subscripts]
                    larrows(x, y, x1, y1, type = "closed", length = 0.1,
                            angle = 15, ...)
                    lpoints(x, y,
                            pch = trellis.par.get("superpose.symbol")$pch[2],
                            col = trellis.par.get("superpose.symbol")$col[2])
                    lpoints(x1, y1,
                            pch = trellis.par.get("superpose.symbol")$pch[1],
                            col = trellis.par.get("superpose.symbol")$col[1])
                    lpoints(ff[2], ff[1], 
                            pch = trellis.par.get("superpose.symbol")$pch[3],
                            col = trellis.par.get("superpose.symbol")$col[3])
                    ltext(fclow[,2], fclow[,1], row.names(fclow),
                          adj = c(0.5, 1.7))
                    ltext(fchigh[,2], fchigh[,1], row.names(fchigh),
                          adj = c(0.5, -0.6))
                  },
                  key = list(space = "top", columns = 3,
                             text = list(c("Mixed model", "Within-group", "Population")),
                             points = list(col = trellis.par.get("superpose.symbol")$col[1:3],
                                           pch = trellis.par.get("superpose.symbol")$pch[1:3]))
     )))

```

Here is a comparison of the random-day effect from `lme` versus a subject-wise linear model. They are not the same.

```{r, echo=FALSE}
print(xyplot(Reaction ~ Days | Subject, sleepstudy, aspect = "xy",
             layout = c(9,2), type = c("g", "p", "r"),
             coef.list = df[,3:4],
             panel = function(..., coef.list) {
               panel.xyplot(...)
               panel.abline(as.numeric(coef.list[packet.number(),]),
                            col.line = trellis.par.get("superpose.line")$col[2],
                            lty = trellis.par.get("superpose.line")$lty[2]
               )
               panel.abline(fixef(lme.3),
                            col.line = trellis.par.get("superpose.line")$col[4],
                            lty = trellis.par.get("superpose.line")$lty[4]
               )
             },
             index.cond = function(x,y) coef(lm(y ~ x))[1],
             xlab = "Days of sleep deprivation",
             ylab = "Average reaction time (ms)",
             key = list(space = "top", columns = 3,
                        text = list(c("Within-subject", "Mixed model", "Population")),
                        lines = list(col = trellis.par.get("superpose.line")$col[c(2:1,4)],
                                     lty = trellis.par.get("superpose.line")$lty[c(2:1,4)]))))
```



## Bibliographic Notes
Most of the examples in this chapter are from the documentation of the __lme4__ package [@lme4]. 
For a more theoretical view see @weiss2005modeling or @searle2009variance.
As usual, a hands on view can be found in @venables2013modern.

## Practice Yourself
