# Unsupervised Learning {#unsupervised}

This chapter deals with machine learning problems which are unsupervised.
This means the machine has access to a set of inputs, $x$, but the desired outcome, $y$ is not available. 
Clearly, learning a relation between inputs and outcomes is impossible, but there are still a lot of problems of interest. 
In particular, we may want to find a compact representation of the inputs, be it for visualization of further processing. This is the problem of _dimensionality reduction_.
For the same reasons we may want to group similar inputs. This is the problem of _clustering_.

In the statistical terminology, and with some exceptions, this chapter can be thought of as multivariate __exploratory__ statistics.
For multivariate __inference__, see Chapter \@ref(multivariate).



## Dimensionality Reduction {#dim-reduce}

```{example, label='bmi'}
Consider the heights and weights of a sample of individuals. 
The data may seemingly reside in $2$ dimensions but given the height, we have a pretty good guess of a persons weight, and vice versa. 
We can thus state that heights and weights are not really two dimensional, but roughly lay on a $1$ dimensional subspace of $\mathbb{R}^2$. 
```

```{example, label='iq'}
Consider the correctness of the answers to a questionnaire with $p$ questions. 
The data may seemingly reside in a $p$ dimensional space, but assuming there is a thing as ``skill'', then given the correctness of a person's reply to a subset of questions, we have a good idea how he scores on the rest. 
Put differently, we don't really need a $200$ question questionnaire-- $100$ is more than enough.
If skill is indeed a one dimensional quality, then the questionnaire data should organize around a single line in the $p$ dimensional cube. 
```


```{example, label='blind-signal'}
Consider $n$ microphones recording an individual. 
The digitized recording consists of $p$ samples. 
Are the recordings really a shapeless cloud of $n$ points in $\mathbb{R}^p$?
Since they all record the same sound, one would expect the $n$ $p$-dimensional points to arrange around the source sound bit: a point in $\mathbb{R}^p$.
If microphones have different distances to the source, volumes may differ. 
We would thus expect the $n$ points to arrange about a line that ends at the source. 	
```

### Principal Component Analysis {#pca}

_Principal Component Analysis_ (PCA) is such a basic technique, it has been rediscovered and renamed independently in many fields. 
It can be found under the names of 
Discrete Karhunen–Loève Transform; Hotteling Transform; Proper Orthogonal Decomposition; Eckart–Young  Theorem; Schmidt–Mirsky Theorem; Empirical Orthogonal Functions; Empirical Eigenfunction Decomposition; Empirical Component Analysis; Quasi-Harmonic Modes; Spectral Decomposition; Empirical Modal Analysis, and possibly more^[http://en.wikipedia.org/wiki/Principal_component_analysis].
The many names are quite interesting as they offer an insight into the different problems that led to PCA's (re)discovery.


Return to the BMI problem in Example \@ref(ex:bmi).
Assume you wish to give each individual a "size score", that is a __linear__ combination of height and weight: PCA does just that. 
It returns the linear combination that has the largest variability, i.e., the combination which best distinguishes between individuals. 

The variance maximizing motivation above was the one that guided @hotelling1933analysis.
But $30$ years before him, @pearson1901liii derived the same procedure with a different motivation in mind. 
Pearson was also trying to give each individual a score. 
He did not care about variance maximization, however. 
He simply wanted a small set of coordinates in some (linear) space that approximates the original data well. 
Before we proceed, we give an example to fix ideas.
Consider the crime rate data in `USArrests`, which encodes reported murder events, assaults, rapes, and the urban population of each americam state. 

```{r}
head(USArrests)
```

Following Hotelling's motivation, we may want to given each state a "crimilality score". 
PCA returns the sequence of $1,\dots,4$ scores that best separate between states. 

```{r, cache=TRUE}
USArrests.1 <- USArrests[,-3] %>% scale # note the scaling, which is required by some
pca.1 <- prcomp(USArrests.1, scale = TRUE)
pca.1
```

Things to note:

- Duistinguishing between states, i.e., finding the variance maximizing scores, should be indifferent to the __average__ of each variable. 
We also don't want the score to be sensitive to the measurement __scale__. 
We thus perform PCA in the z-score scale of each variable, obtained with the `scale` function.
- PCA is performed with the `prcomp` function. 
It returns the contribution (weight) of the original variables, to the new crimeness score.  
These weights are called the _loadings_.
- The number of possible scores, is the same as the number of original variables in the data. 
- The new scores are called the _principal components_, labeled `PC1`,...,`PC4` in our output.
- The loadings on PC1 tell us that the best separation between states is along the average crime rate. 
Why is this? 
Because all the $3$ crime variables have a similar loading on PC1.
- The other PCs are slightly harder to interpret, but it is an interesting excercise.

__If we now represent each state, not with its original $4$ variables, but only with the first $2$ PCs (for example), we have reduced the dimensionality of the data.__




### Preliminaries

Before presenting methods other than PCA, we need some terminology.

- __Variable__: 
A.k.a. _dimension_, or _feature_, or _column_ for reasons that will be obvious in the next item.
- __Data__: 
A.k.a. _sample_, _observations_. 
Will typically consist of $n$, $p$ dimensional vectors.
We typically denote the data as a $n\times p$ matrix $X$. 
- __Manifold__: 
A generalization of a linear space, which is regular enough so that, __locally__, it has all the properties of a linear space. 
We will denote an arbitrary manifold by $\mathcal{M}$, and by $\mathcal{M}_q$ a $q$ dimensional^[You are probably used to thinking of the __dimension__ of linear spaces. We will not rigorously define what is the dimension of a manifold, but you may think of it as the number of free coordinates needed to navigate along the manifold.] manifold.
- __Embedding__: 
Informally speaking: a ``shape preserving'' mapping of a space into another. 
- __Linear Embedding__: 
An embedding done via a linear operation (thus representable by a matrix). 
- __Generative Model__: 
Known to statisticians as the __sampling distribution__. 
The assumed stochastic process that generated the observed $X$. 

There are many motivations for dimensionality reduction:

1. __Scoring__: 
Give each observation an interpretable, simple score (Hotelling's motivation).
1. __Latent structure__: 
Recover unobservables from indirect measurements. 
E.g: Blind signal reconstruction, CT scan, cryo-electron microscopy, etc. 
1. __Signal to Noise__: 
Denoise measurements before further processing like clustering, supervised learning, etc. 
1. __Compression__: 
Save on RAM ,CPU, and communication when operating on a lower dimensional representation of the data. 
	


### Latent Variable Approaches
All generative approaches to dimensionality reduction will include some unobserved set of variables, which we can try to recover from the observable $X$. 
The unobservable variables will typically have a lower dimension than the observables, thus, dimension is reduced. 
We start with the simplest case of linear Factor Analysis. 

#### Factor Analysis (FA)

FA originates from the psychometric literature. 
We thus revisit the IQ (actually g-factor^[https://en.wikipedia.org/wiki/G_factor_(psychometrics)]) Example~\@ref(ex:iq):

```{example}
Assume $n$ respondents answer $p$ quantitative questions: $x_i \in \mathbb{R}^p, i=1,\dots,n$. 
Also assume, their responses are some linear function $A \in \mathbb{R}^p$ of a single personality attribute, $s_i$. 
We can think of $s_i$ as the subject's ``intelligence''.
We thus have 
\begin{align}
	x_i = A s_i + \varepsilon_i
\end{align}
And in matrix notation, for $q<p$ latent attributes:
\begin{align}
	X = S A+\varepsilon,
	(\#eq:factor)
\end{align}
where $A$ is the $q \times p$ matrix of factor loadings, and $S$ the $n \times q$ matrix of latent personality traits. 
In our particular example where $q=1$, the problem is to recover the unobservable intelligence scores, $s_1,\dots,s_n$, from the observed answers $X$.	
```



We may try to estimate $S A$ by assuming some distribution on $S$ and $\varepsilon$ and apply maximum likelihood.
Under standard assumptions on the distribution of $S$ and $\varepsilon$, recovering  $S$ from $\widehat{S A }$ is still impossible as there are infinitely many such solutions.
In the statistical parlance we say the problem is _non identifiable_, and in the applied mathematics parlance we say the problem is _ill posed_.
To see this, consider an orthogonal _rotation_ matrix $R$ ($R' R=I$). 
For each such $R$: $ S A = A R' R S = S^* A^* $.
While both solve Eq.\@ref(eq:factor), $A$ and $A^*$ may have very different interpretations. 
This is why many researchers find FA an unsatisfactory inference tool.

```{remark}
The non-uniqueness (non-identifiability) of the FA solution under variable rotation is never mentioned in the PCA context. 
Why is this?
This is because the methods solve different problems. 
The reason the solution to PCA is well defined is that PCA does not seek a single $S$ but rather a __sequence__ of $S_q$ with dimensions growing from $q=1$ to $q=p$. 
```

```{remark}
In classical FA in Eq.\@ref(eq:factor) is clearly an embedding to a linear space. 
The one spanned by $S$. 
Under the classical probabilistic assumptions on $S$ and $\varepsilon$ the embedding itself is also linear, and is sometimes solved with PCA. 
Being a generative model, there is no restriction for the embedding to be linear, and there certainly exists sets of assumptions for which the FA embedding is non linear. 
```

The FA terminology is slightly different than PCA:

- __Factors__: 
The unobserved attributes $S$. 
Not to be confused with the _principal components_ in the context of PCA.
- __Loading__:
The $A$ matrix; the contribution of each factor to the observed $X$.
- __Rotation__:	
An arbitrary orthogonal re-combination of the factors, $S$, and loadings, $A$, which changes the interpretation of the result.


The FA literature does offer several heuristics to ``fix'' the solution of the FA. 
These are known as _rotations_, and go under the names of _Varimax_, _Quartimax_, _Equimax_, _Oblimin_, _Promax_, and possibly others. 


#### Independent Component Analysis (ICA)
Like FA, _independent compoent analysis_ (ICA) is a family of latent space models, thus, a _meta-method_.
It assumes data is generated as some function of the latent variables $S$. 
In many cases this function is assumed to be linear in $S$ so that ICA is compared, if not confused, with PCA and even more so with FA. 

The fundamental idea of ICA is that $S$ has a joint distribution of __non-Gaussian__, __independent__ variables. 
This independence assumption, solves the the non-uniquness of $S$ in FA.

Being a generative model, estimation of $S$ can then be done using maximum likelihood, or other estimation principles. 

ICA is a popular technique in signal processing, where $A$ is actually the signal, such as sound in Example \@ref(ex:blind-signal).
Recovering $A$ is thus recovering the original signals mixing in the recorded $X$. 





### Purely Algorithmic Approaches

We now discuss dimensionality reduction approaches that are not stated via their generative model, but rather, directly as an algorithm.
This does not mean that they cannot be cast via their generative model, but rather they were not motivated as such.


#### Multidimensional Scaling (MDS)
MDS can be thought of as a variation on PCA, that begins with a distance graph^[The term Graph is typically used in this context instead of Network. But a graph allows only yes/no relations, while a network, which is a weighted graph, allows a continuous measure of similarity (or dissimilarity). _Network_ is thus more appropriate than _graph_.]}.

MDS aims at embedding a graph of distances, while preserving the original distances.
Basic results in graph/network theory [@graham1988isometric] suggest that the geometry of a graph cannot be preserved when embedding it into lower dimensions. 
The different types of MDSs, such as _Classical MDS_, and _Sammon Mappings_, differ in the _stress function_ penalizing for geometric distortion.


#### Local Multidimensional Scaling (Local MDS)

```{example, label='non-euclidean'}
Consider data of coordinates on the globe. 
At short distances, constructing a dissimilarity graph with Euclidean distances will capture the true distance between points. 
At long distances, however, the Euclidean distances as grossly inappropriate. 
A more extreme example is coordinates on the brain's cerebral cortex.
Being a highly folded surface, the Euclidean distance between points is far from the true geodesic distances along the cortex's surface^[Then again, it is possible that the true distances are the white matter fibers connecting going within the cortex, in which case, Euclidean distances are more appropriate than geodesic distances. We put that aside for now.].
```

Local MDS is aimed at solving the case where we don't know how to properly measure distances. 
It is an algorithm that compounds both the construction of the dissimilarity graph, and the embedding. 
The solution of local MDS, as the name suggests, rests on the computation of _local_ distances, where the Euclidean assumption may still be plausible, and then aggregate many such local distances, before calling upon regular MDS for the embedding.

Because local MDS ends with a regular MDS, it can be seen as a non-linear embedding into a linear $\mathcal{M}$. 

Local MDS is not popular.
Why is this? 
Because it makes no sense: 
If we believe the points reside in a non-Euclidean space, thus motivating the use of geodesic distances, why would we want to wrap up with regular MDS, which embeds in a linear space?!


#### Isometric Feature Mapping (IsoMap) {#isomap}
Like localMDS, only that the embedding, and not only the computation of the distances is local.


#### Local Linear Embedding (LLE)
Very similar to IsoMap \@ref(isomap).


#### Kernel PCA
TODO

#### Simplified Component Technique LASSO (SCoTLASS)
TODO

#### Sparse Principal Component Analysis (sPCA)
TODO

#### Sparse kernel principal component analysis (skPCA)
TODO




### Dimensionality Reduction in R

#### PCA

We already saw the basics of PCA in \@ref(pca).
The fitting is done with the `procomp` function.
The _bi-plot_ is a useful way to visualize the output of PCA.

```{r}
library(devtools)
# install_github("vqv/ggbiplot")
ggbiplot::ggbiplot(pca.1) 
```

Things to note:

- The bi-plot plots each data point along its PCs.
- We used the `ggbiplot` function from the __ggbiplot__ (available from github, but not from CRAN), because it has a nicer output than `stats::biplot`. 
- The bi-plot also plots the loadings as arrows. The coordinates of the arrows belong to the weight of each of the original variables in each PC. 
For example, the x-value of each arrow is the loadings on the first PC (on the x-axis).
Since the weights of Murder, Assult, and Rape are almost the same, and larger then UrbanPop, we conclude that PC1 captures the avarage crime rate in each state.

The _scree plot_ depicts the quality of the approximation of $X$ as $q$ grows. 
This is depicted using the proportion of variability in $X$ that is removed by each added PC.
It is customary to choose $q$ as the first PC that has a relative low contribution to the approximation of $X$.

```{r}
ggbiplot::ggscreeplot(pca.1)
```

See how the first PC captures the variability in the Assault levels and Murder levels, with a single score.

```{r, echo=FALSE}
USArrests.1 <- USArrests[,-3] %>% scale
load <- pca.1$rotation
slope <- load[2, ]/load[1, ]
mn <- apply(USArrests.1, 2, mean)
intcpt <- mn[2] - (slope * mn[1])

# scatter plot with the two new axes added
USArrests.2 <- USArrests[,1:2] %>%  scale
xlim <- range(USArrests.2)  # overall min, max
plot(USArrests.2, xlim = xlim, ylim = xlim, pch = 16, col = "purple")  # both axes same length
abline(intcpt[1], slope[1], lwd = 2)  # first component solid line
abline(intcpt[2], slope[2], lwd = 2, lty = 2)  # second component dashed
legend("right", legend = c("PC 1", "PC 2"), lty = c(1, 2), lwd = 2, cex = 1)

# projections of points onto PCA 1
y1 <- intcpt[1] + slope[1] * USArrests.2[, 1]
x1 <- (USArrests.1[, 2] - intcpt[1])/slope[1]
y2 <- (y1 + USArrests.1[, 2])/2
x2 <- (x1 + USArrests.1[, 1])/2
segments(USArrests.1[, 1], USArrests.1[, 2], x2, y2, lwd = 2, col = "purple")
```

More implementations of PCA:
```{r many PCA implementations, eval=FALSE}
# FAST solutions:
gmodels::fast.prcomp()

# More detail in output:
FactoMineR::PCA()

# For flexibility in algorithms and visualization:
ade4::dudi.pca()

# Another one...
amap::acp()
```


#### FA


```{r FA, cache=TRUE}
fa.1 <- psych::principal(USArrests.1, nfactors = 2, rotate = "none")
fa.1
biplot(fa.1, labels =  rownames(USArrests.1)) 

# Numeric comparison with PCA:
fa.1$loadings
pca.1$rotation
```

Things to note:

- We perform FA with the `psych::principal` function.
- The first factor (`fa.1$loadings`) has different weights than the firts PC (`pca.1$rotation`) because of normalization. They are the same, however, in that the first PC, and the first factor, capture average crime levels.

Graphical model fans will like the following plot, where the contribution of each variable to each factor is encoded in the width of the arrow. 

```{r, cache=TRUE}
# Graph comparison: loadings encoded in colors
qgraph::qgraph(fa.1)
```

Let's add a rotaion (Varimax), and note that the rotation has indeed changed the loadings of the variables, thus the interpretation of the factors. 

```{r varimax, cache=TRUE}
fa.2 <- psych::principal(USArrests.1, nfactors = 2, rotate = "varimax")

fa.2$loadings
```




#### ICA

```{r ICA, cache=TRUE}
ica.1 <- fastICA::fastICA(USArrests.1, n.com=2) # Also performs projection pursuit

plot(ica.1$S)
abline(h=0, v=0, lty=2)
text(ica.1$S, pos = 4, labels = rownames(USArrests.1))

# Compare with two PCA (first two PCs):
arrows(x0 = ica.1$S[,1], y0 = ica.1$S[,2], x1 = pca.1$x[,2], y1 = pca.1$x[,1], col='red', pch=19, cex=0.5)
```

Things to note:

- ICA is fitted with `fastICA::fastICA`.
- The ICA componetns are very different than the PCA components. 




#### MDS

Classical MDS, also compared with PCA.
```{r MDS, results='hold', cache=TRUE}
# We first need a dissimarity matrix/graph:
state.disimilarity <- dist(USArrests.1)

mds.1 <- cmdscale(state.disimilarity)

plot(mds.1, pch = 19)
abline(h=0, v=0, lty=2)
USArrests.2 <- USArrests[,1:2] %>%  scale
text(mds.1, pos = 4, labels = rownames(USArrests.2), col = 'tomato')

# Compare with two PCA (first two PCs):
points(pca.1$x[,1:2], col='red', pch=19, cex=0.5)
```

Things to note:

- For MDS, we firts compute a dissimilarity graph with `dist`, and then learn the embedding with `cmdscale`.
- As previously stated, the embedding of PCA is the same as classical MDS with Euclidean distances. 
- See the `cluster::daisy` function for more dissimilarity measures.


Let's try other strain functions for MDS, like Sammon's strain, and compare it with the PCs.
```{r Sammon MDS}
mds.2 <- MASS::sammon(state.disimilarity, trace = FALSE)
plot(mds.2$points, pch = 19)
abline(h=0, v=0, lty=2)
text(mds.2$points, pos = 4, labels = rownames(USArrests.2))

# Compare with two PCA (first two PCs):
arrows(x0 = mds.2$points[,1], y0 = mds.2$points[,2], x1 = pca.1$x[,1], y1 = pca.1$x[,2], col='red', pch=19, cex=0.5)
```

Things to note:

- `MASS::sammon` does the fitting.
- The embedding returned by the Sammon strain is different than that of the first two PCs.






#### Sparse PCA

```{r sPCA}
# Compute similarity graph
state.similarity <- MASS::cov.rob(USArrests.1)$cov

spca1 <- elasticnet::spca(state.similarity, K=2, type="Gram", sparse="penalty", trace=FALSE, para=c(0.06,0.16))
spca1$loadings
```



#### Kernel PCA
```{r kPCA, eval=FALSE}
kernlab::kpca()
```




## Clustering {#cluster}





### Clustering in R





## Bibliographic Notes
For more on PCA see my [Dimensionality Reduction Class Notes](https://github.com/johnros/dim_reduce/blob/master/dim_reduce.pdf) and references therein.


