# Plotting {#plotting}

Whether you are doing EDA, or preparing your results for publication, you need plots.
R has many plotting mechanisms, allowing the user a tremendous amount of flexibility, while abstracting away a lot of the tedious details. 
To be concrete, many of the plots in R are simply impossible to produce with Excel, SPSS, or SAS, and would take a tremendous amount of work to produce with Python, Java and lower level programming languages. 

In this text, we will focus on two plotting packages. 
The basic __graphics__ package, distributed with the base R distribution, and the __ggplot2__ package. 

Before going into the details of the plotting packages, we start with some high-level philosophy.
The __graphics__ package originates from the main-frame days. 
Computers had no graphical interface, and the output of the plot was immediately sent to a printer. 
For this reason, once a plot has been produced with the __graphics__ package, it cannot be queryied nor changed, except for further additions.

The philosophy of R is that __everyting is an object__.
The __graphics__ package does not adhere to this philosophy, and indeed it was soon augmented with the __grid__ package [@Rlanguage], that treats plots as objects. 
__grid__ is a low level graphics interface, and users may be more familiar with the __lattice__ package built upon it [@lattice].

__lattice__ is very powerful, but soon enough, it was overtaken in popularity by the __ggplot2__ package [@ggplot2].
__ggplot2__ was the PhD project of [Hadley Wickham](http://hadley.nz/), a name to remember...
Two fundamental ideas underlay __ggplot2__: (i) everything is an object, and (ii), plots can be described by a small set of building blocks. 
The building blocks in __ggplot2__ are the ones stated by @wilkinson2006grammar.
The objects and grammar of __ggplot2__ have later evolved to allow more complicated plotting and in particular, interactive plotting, in other packages. 

Interactive plotting is a very important feature for EDA, and for reporting. 
The major leap in interactive plotting was made possible by the advancement of web technologies, such as JavaScript.
Why is this? 
Because an interactive plot, or report, can be seen as a web-site. 
Building upon the capabilities of JavaScript, and your web browser, to provide the interactivity, greatly facilitates the development of such plots, as the programmer can reply on the web-browsers capabilities for interactivity. 

One of the latest contributions to interactive plotting, is the __Shiny__ framework by RStudio [@shiny].
__Shiny__, unlike other interactive plotting systems, is not a static web-site. 
It is a web-server, that can query R, with all its facilities.


## The graphics System

The R code from the Basics Chapter \@ref(basics) is a demonstration of the __graphics__ package and system. 
We make a quick review of the basics.

### Using Existing Plotting Functions

#### Scatter Plot

A simple scatter plot.
```{r}
attach(trees)
plot(Girth ~ Height)
```

Various types of plots.

```{r}
par(mfrow=c(2,3))
plot(Girth, type='h', main="type='h'") 
plot(Girth, type='o', main="type='o'") 
plot(Girth, type='l', main="type='l'")
plot(Girth, type='s', main="type='s'")
plot(Girth, type='b', main="type='b'")
plot(Girth, type='p', main="type='p'")
par(mfrow=c(1,1))
```

Things to note:
TODO


Controll the plotting charachters
```{r}
plot(Girth, pch='+', cex=3)
```

Control the lines' appearance with the `lty` argument.

```{r}
par(mfrow=c(2,3))
plot(Girth, type='l', lty=1, lwd=2)
plot(Girth, type='l', lty=2, lwd=2)
plot(Girth, type='l', lty=3, lwd=2)
plot(Girth, type='l', lty=4, lwd=2)
plot(Girth, type='l', lty=5, lwd=2)
plot(Girth, type='l', lty=6, lwd=2)
par(mfrow=c(1,1))
```

Add line by slope and intercept with `abline`.

```{r}
plot(Girth)

abline(v=14, col='red') 
abline(h=9, lty=4,lwd=4, col='pink') 
abline(0,1 )
```

Things to note:
TODO
- Locations may be added interactively using the `locator` function.

```{r}
plot(Girth)
points(x=1:30, y=rep(12,30), cex=0.5, col='darkblue')

lines(x=rep(c(5,10), 7), y=7:20, lty=2 )
lines(x=rep(c(5,10), 7)+2, y=7:20, lty=2 )
lines(x=rep(c(5,10), 7)+4, y=7:20, lty=2 , col='darkgreen')
lines(x=rep(c(5,10), 7)+6, y=7:20, lty=4 , col='brown', lwd=4)
```

Things to note:
TODO


Add line segments.

```{r}
plot(Girth)

segments(x0=rep(c(5,10), 7), y0=7:20, x1=rep(c(5,10), 7)+2, y1=(7:20)+2 )

arrows(x0=13,y0=16,x1=16,y1=17, )

rect(xleft=10, ybottom=12,  xright=12, ytop=16)

polygon(x=c(10,11,12,11.5,10.5), y=c(9,9.5,10,10.5,9.8), col='grey')

title(main='This plot makes no sense', sub='Or does it?')

mtext('Printing in the margins', side=2)
mtext(expression(alpha==log(f[i])), side=4)
```

Things to note:
TODO
- For more information for mathematical annotation see `?plotmath`.

Add a legend.
```{r}
plot(Girth, pch='G',ylim=c(8,77), xlab='Tree number', ylab='', type='b', col='blue')
points(Volume, pch='V', type='b', col='red')
legend(x=2, y=70, legend=c('Girth', 'Volume'), pch=c('G','V'), col=c('blue','red'), bg='grey')
```


Adjusting Axes.
```{r}
plot(Girth, xlim=c(0,15), ylim=c(8,12))
```

Use `layout` for complicated plot layouts.

```{r}
A<-matrix(c(1,1,2,3,4,4,5,6), byrow=TRUE, ncol=2)
layout(A,heights=c(1/14,6/14,1/14,6/14))

oma.saved <- par("oma")
par(oma = rep.int(0, 4))
par(oma = oma.saved)
o.par <- par(mar = rep.int(0, 4))
for (i in seq_len(6)) {
	plot.new()
	box()
	text(0.5, 0.5, paste('Box no.',i), cex=3)
}
```


Always detach.
```{r}
detach(trees)
```

#### Univariate Plotting

```{r}
attach(PlantGrowth)
head(PlantGrowth)
```



```{r}
plot(group)
pie(table(group))
```


```{r}
plot(weight)
```

```{r}
plot(ecdf(weight))
```


```{r}
hist(weight)
rug(weight)
```

```{r}
plot(density(weight))
rug(weight)
```


```{r}
fact.weight<- cut(weight, breaks=2, labels=c('Light', 'Heavy'))
plot(table(group, fact.weight))
```


```{r}
dotchart(table(fact.weight,group), pch=c(1,4), col=c('seagreen','grey'))
dotchart(table(group,fact.weight), col=c('red','darkred','blue'))
```


```{r}
boxplot(weight)
```


```{r}
par(mfrow=c(3,1))
hist(weight[group=='trt1'], ylim=c(0,3),xlim=c(3,6))
hist(weight[group=='trt2'], ylim=c(0,3),xlim=c(3,6))
hist(weight[group=='ctrl'], ylim=c(0,3),xlim=c(3,6))
par(mfrow=c(1,1))
```

```{r}
boxplot(weight ~ group)
```

```{r}
boxplot(weight~group, col=c('seagreen2','lightgrey','darkred'))
```



dens.trt1<- density(weight[group=='trt1'])
dens.trt2<- density(weight[group=='trt2'])
dens.ctrl<- density(weight[group=='ctrl'])
plot(dens.trt1, xlim=c(2,8), ylim=c(0,1), col='seagreen')
lines(dens.trt2, col='grey')
lines(dens.ctrl, col='darkred')
legend(x=2, y=1, legend=c('trt1','trt2','ctrl'), col=c('seagreen', 'grey', 'darkred'),lty=1)

par(mfrow=c(3,1))
hist(weight[group=='trt1'], probability=TRUE, xlim=c(2,8))
lines(dens.trt1)
rug(weight[group=='trt1'])
hist(weight[group=='trt2'], probability=TRUE, xlim=c(2,8))
lines(dens.trt2)
rug(weight[group=='trt2'])
hist(weight[group=='ctrl'], probability=TRUE, xlim=c(2,8))
lines(dens.ctrl)
rug(weight[group=='ctrl'])
par(mfrow=c(1,1))


require(lattice)
dotplot(weight~group)

detach(PlantGrowth)
#############################################################
#--------------- Multivariate Display-----------------------#
#############################################################
head(trees)
attach(trees)

# Mosaic plot for categorical variables
fac.height<- factor(cut(Height,2), labels=c('Short','Tall'))
fac.Volume<- factor(cut(Volume, 3), labels=c('Small', 'Medium', 'Large'))
fac.Girth<- factor(cut(Girth, 2), labels=c('Fat', 'Thin'))

# Two at a time:
table(fac.Volume, fac.height)
mosaicplot(table(fac.Volume, fac.height))

# Three at a time:
table(fac.Volume, fac.height, fac.Girth)
ftable(fac.Volume, fac.height, fac.Girth)
mosaicplot(table(fac.height, fac.Volume , fac.Girth))


ftable(Titanic)
mosaicplot(Titanic)


# Scatter plots with our first OLS line!
plot(Volume~Girth)
abline(lm(Volume ~ Girth))
(lm.coefs<- round(coef(lm(Volume~Girth)),2))  
text(x=12, y=70, paste('Slope=', lm.coefs[2],'\n Intercept=', lm.coefs[1] ))
rug(Girth, side=1)
rug(Volume, side=2)

plot(Volume~Girth, col=fac.height)
legend(x=9, y=71, legend=c('Short','Tall'), col=c(1,2), pch=1)

# Printer friendly version
pchs<- as.numeric(fac.height) 
plot(Volume~Girth, pch=pchs)
legend(x=9, y=71, legend=c('Short','Tall'), pch=unique(pchs))
abline(lm(Volume[fac.height=='Short']~Girth[fac.height=='Short']))
abline(lm(Volume[fac.height!='Short']~Girth[fac.height!='Short']), lty=2)

#Interaction with the plot
plot(Volume~Girth)
identify(y=Volume, x=Girth) # Press ESC to exit interactive mode

# Scatter plot matrix:
plot(trees) #this actually calls: 
pairs(trees)

# A fancier version of the scatter plot matrix:
require(car)
scatterplot.matrix(trees)

# A conditional scatterplot:
coplot(Volume ~ Girth | fac.height)

# 3D scatter plot with fitted plane
require(scatterplot3d)
s3d<- scatterplot3d(x=Girth, y=Height , z=Volume , type='h', highlight.3d=T)

(lm.2<- lm(Volume ~  Girth + Height))
s3d$plane3d(lm.2, lty.box='solid')

```




### Generating a New Plot
```{r}
------------ Starting a new plot-------------#
plot.new()

# Margin sizes in inches:
par(mai=c(2, 2, 1, 1))
# Margin sizes in lines of text:
par(mar=c(4, 4, 2, 2))
# Plot size in inches:
par(pin=c(5, 4))

# Set axis scals:
plot.window(xlim=c(0,1), ylim=c(10,20))

plot.new()

plot.window(xlim=c(0,1), ylim=c(10,20))

axis(side=1)
axis(side=2)
axis(side=3)
axis(side=4, pos=1.1)

plot.new()
plot.window(xlim=c(1,4), ylim=c(10,20))
axis(1, at=1:4, labels=c('A', 'B', 'C' ,'D'), col='dark orange', col.axis='red', font.axis= 2)
title(main='Main', xlab='X', ylab='Y', font.main= 2, font.sub=3)
title(sub='Sub', col.sub='dark green', cex.sub=10)
box()

# A line graph:
x = 1995:2005
y = c(81.1, 83.1, 84.3, 85.2, 85.4, 86.5, 88.3, 88.6, 90.8, 91.1, 91.3)
plot.new()
plot.window(xlim = range(x), ylim = range(y))
abline(h = -4:4, v = -4:4, col = "lightgrey")
lines(x, y, lwd = 2)
title(main = "A Line Graph Example",
		xlab = "Time",
		ylab = "Quality of R Graphics")
axis(1)
axis(2)
box()

# Rosette:
n = 17
theta = seq(0, 2 * pi, length = n + 1)[1:n]
x = sin(theta)
y = cos(theta)
(v1 = rep(1:n, n))
(v2 = rep(1:n, rep(n, n)))
plot.new()
plot.window(xlim = c(-1, 1), ylim = c(-1, 1), asp = 1)
segments(x[v1], y[v1], x[v2], y[v2])
box()

# Curve Envelope:
x1 = seq(0, 1, length = 20)
y1 = rep(0, 20)
x2 = rep(0, 20)
y2 = seq(0.75, 0, length = 20)
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 0.75), asp = 1)
segments(x1, y1, x2, y2)
box(col = "grey")

# Arrows:
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
arrows(.05, .075, .45, .9, code = 1)
arrows(.55, .9, .95, .075, code = 2)
arrows(.1, 0, .9, 0, code = 3)
text(.5, 1, "A", cex = 1.5)
text(0, 0, "B", cex = 1.5)
text(1, 0, "C", cex = 1.5)

# Arrows as error bars:
x = 1:10
y = runif(10) + rep(c(5, 6.5), c(5, 5))
yl = y - 0.25 - runif(10)/3
yu = y + 0.25 + runif(10)/3
plot.new()
plot.window(xlim = c(0.5, 10.5), ylim = range(yl, yu))
arrows(x, yl, x, yu, code = 3, angle = 90, length = .125)
points(x, y, pch = 19, cex = 1.5)
axis(1, at = 1:10, labels = LETTERS[1:10])
axis(2, las = 1)
box()

# Manual boxplot:
plot.new()
plot.window(xlim = c(0, 5), ylim = c(0, 10))
rect(0:4, 0, 1:5, c(7, 8, 4, 3), col = "lightblue")
axis(1)
axis(2, las = 1)

# Spiral Squares:
plot.new()
plot.window(xlim = c(-1, 1), ylim = c(-1, 1), asp = 1)
x = c(-1, 1, 1, -1)
y = c( 1, 1, -1, -1)
polygon(x, y, col = "cornsilk")
vertex1 = c(1, 2, 3, 4)
vertex2 = c(2, 3, 4, 1)
for(i in 1:50) {
	x = 0.9 * x[vertex1] + 0.1 * x[vertex2]
	y = 0.9 * y[vertex1] + 0.1 * y[vertex2]
	polygon(x, y, col = "cornsilk")
}

# Drawing text:
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
abline(h = c(.2, .5, .8), v = c(.5, .2, .8), col = "lightgrey")
text(0.5, 0.5, "srt = 45, adj = c(.5, .5)", srt=45, adj=c(.5, .5))
text(0.5, 0.8, "adj = c(0, .5)", adj = c(0, .5))
text(0.5, 0.2, "adj = c(1, .5)", adj = c(1, .5))
text(0.2, 0.5, "adj = c(1, 1)", adj = c(1, 1))
text(0.8, 0.5, "adj = c(0, 0)", adj = c(0, 0))
axis(1)
axis(2, las = 1)
box()

# Circles (Just a polygon...):
R = 1
xc = 0
yc = 0
n = 72
t = seq(0, 2 * pi, length = n)[1:(n-1)]
(x = xc + R * cos(t))
(y = yc + R * sin(t))
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
polygon(x, y, col = "lightblue", border = "navyblue")

# Ellipse:
a = 4
b = 2
xc = 0
yc = 0
n = 72
t = seq(0, 2 * pi, length = n)[1:(n-1)]
(x = xc + a * cos(t))
(y = yc + b * sin(t))
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
polygon(x, y, col = "lightblue")

# Rotated Ellipse:
a = 4
b = 2
xc = 0
yc = 0
n = 72
theta = 45 * (pi / 180)
t = seq(0, 2 * pi, length = n)[1:(n-1)]
x = xc + a * cos(t) * cos(theta) - b * sin(t) * sin(theta)
y = yc + a * cos(t) * sin(theta) + b * sin(t) * cos(theta)
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
polygon(x, y, col = "lightblue")

# Bivariate normal quantile:
n = 72
rho = 0.5
d = acos(rho)
t = seq(0, 2 * pi, length = n)[1:(n-1)]
plot.new()
plot.window(xlim = c(-1, 1), ylim = c(-1, 1), asp = 1)
rect(-1, -1, 1, 1)
polygon(cos(t + d), y = cos(t))
segments(-1, 0, 1, 0, lty = "13")
segments(0, -1, 0, 1, lty = "13")
axis(1)
axis(2, las = 1)

# Spiral:
k = 5
n = k * 72
theta = seq(0, k * 2 * pi, length = n)
R = .98^(1:n - 1)
x = R * cos(theta)
y = R * sin(theta)
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
lines(x, y)

# Querying the device status:
par('din') # Device dimentions in inches
par('fin') # Current figure dimensions in inches
par('pin') # Current plot region dimensions in inches
par('fig') # Normalized Device Coordinates for the figure region
par('plt') # NDC coordinates for the plot region
par('usr') # Extract exact scale limits: xleft, xright, ybottom, ytop.


# Calculate angles:
x = c(0, 0.5, 1.0)
y = c(0.25, 0.5, 0.25)
plot(x, y, type = "l")
(dx = diff(x))
(dy = diff(y))
(pin = par("pin"))
(usr = par("usr"))
(ax = pin[1]/diff(usr[1:2]))
(ay = pin[2]/diff(usr[3:4]))
diff(180 * atan2(ay * dy, ax * dx) / pi) # Degrees of direction change



```




### Adding Elements



### Controlling Plotting Parameters



### Exporting a Plot




## The ggplot2 System


### Generating a New Plot




### Adding Elements



### Controlling Plotting Parameters







## Interactive Graphics






## Shiny


