# Plotting {#plotting}

Whether you are doing EDA, or preparing your results for publication, you need plots.
R has many plotting mechanisms, allowing the user a tremendous amount of flexibility, while abstracting away a lot of the tedious details. 
To be concrete, many of the plots in R are simply impossible to produce with Excel, SPSS, or SAS, and would take a tremendous amount of work to produce with Python, Java and lower level programming languages. 

In this text, we will focus on two plotting packages. 
The basic __graphics__ package, distributed with the base R distribution, and the __ggplot2__ package. 

Before going into the details of the plotting packages, we start with some high-level philosophy.
The __graphics__ package originates from the main-frame days. 
Computers had no graphical interface, and the output of the plot was immediately sent to a printer. 
For this reason, once a plot has been produced with the __graphics__ package, it cannot be queryied nor changed, except for further additions.

The philosophy of R is that __everyting is an object__.
The __graphics__ package does not adhere to this philosophy, and indeed it was soon augmented with the __grid__ package [@Rlanguage], that treats plots as objects. 
__grid__ is a low level graphics interface, and users may be more familiar with the __lattice__ package built upon it [@lattice].

__lattice__ is very powerful, but soon enough, it was overtaken in popularity by the __ggplot2__ package [@ggplot2].
__ggplot2__ was the PhD project of [Hadley Wickham](http://hadley.nz/), a name to remember...
Two fundamental ideas underlay __ggplot2__: (i) everything is an object, and (ii), plots can be described by a small set of building blocks. 
The building blocks in __ggplot2__ are the ones stated by @wilkinson2006grammar.
The objects and grammar of __ggplot2__ have later evolved to allow more complicated plotting and in particular, interactive plotting, in other packages. 

Interactive plotting is a very important feature for EDA, and for reporting. 
The major leap in interactive plotting was made possible by the advancement of web technologies, such as JavaScript.
Why is this? 
Because an interactive plot, or report, can be seen as a web-site. 
Building upon the capabilities of JavaScript, and your web browser, to provide the interactivity, greatly facilitates the development of such plots, as the programmer can reply on the web-browsers capabilities for interactivity. 

One of the latest contributions to interactive plotting, is the __Shiny__ framework by RStudio [@shiny].
__Shiny__, unlike other interactive plotting systems, is not a static web-site. 
It is a web-server, that can query R, with all its facilities.


## The graphics System

The R code from the Basics Chapter \@ref(basics) is a demonstration of the __graphics__ package and system. 
We make a quick review of the basics.

### Using Existing Plotting Functions

#### Scatter Plot

A simple scatter plot.
```{r}
attach(trees)
plot(Girth ~ Height)
```

Various types of plots.

```{r}
par(mfrow=c(2,3))
plot(Girth, type='h', main="type='h'") 
plot(Girth, type='o', main="type='o'") 
plot(Girth, type='l', main="type='l'")
plot(Girth, type='s', main="type='s'")
plot(Girth, type='b', main="type='b'")
plot(Girth, type='p', main="type='p'")
par(mfrow=c(1,1))
```

Things to note:

- The `par` command controls the plotting parameters. `mfrow=c(2,3)` is used to produce a matix of plots with 2 rows and 3 columns.
- The `type` argument controls the type of plot. 
- The `main` argument controls the title.
- See `?plot` and `?par` for more options.

Controll the plotting charachters with the `pch` argument. 
```{r}
plot(Girth, pch='+', cex=3)
```

Control the line's type with `lty` argument, and width with `lwd`.

```{r}
par(mfrow=c(2,3))
plot(Girth, type='l', lty=1, lwd=2)
plot(Girth, type='l', lty=2, lwd=2)
plot(Girth, type='l', lty=3, lwd=2)
plot(Girth, type='l', lty=4, lwd=2)
plot(Girth, type='l', lty=5, lwd=2)
plot(Girth, type='l', lty=6, lwd=2)
par(mfrow=c(1,1))
```

Add line by slope and intercept with `abline`.

```{r}
plot(Girth)
abline(v=14, col='red') # vertical line at 14.
abline(h=9, lty=4,lwd=4, col='pink') # horizontal line at 9.
abline(a = 0, b=1) # linear line with intercept a=0, and slope b=1.
```

```{r}
plot(Girth)
points(x=1:30, y=rep(12,30), cex=0.5, col='darkblue')
lines(x=rep(c(5,10), 7), y=7:20, lty=2 )
lines(x=rep(c(5,10), 7)+2, y=7:20, lty=2 )
lines(x=rep(c(5,10), 7)+4, y=7:20, lty=2 , col='darkgreen')
lines(x=rep(c(5,10), 7)+6, y=7:20, lty=4 , col='brown', lwd=4)
```

Things to note:
- `points` adds points on an existing plot.
- `lines` adds lines on an existing plot.
- `col` controls the color of the element. It takes names or numbers as argument. 
- `cex` controls the scale of the element. Defaults to `cex=1`.


Add other elements.

```{r}
plot(Girth)
segments(x0=rep(c(5,10), 7), y0=7:20, x1=rep(c(5,10), 7)+2, y1=(7:20)+2 )
arrows(x0=13,y0=16,x1=16,y1=17, )
rect(xleft=10, ybottom=12,  xright=12, ytop=16)
polygon(x=c(10,11,12,11.5,10.5), y=c(9,9.5,10,10.5,9.8), col='grey')
title(main='This plot makes no sense', sub='Or does it?')
mtext('Printing in the margins', side=2)
mtext(expression(alpha==log(f[i])), side=4)
```

Things to note:
- The following functions add the elements they are names after: `segments`, `arrows`, `rect`, `polygon`, `title`. 
- `mtext` adds mathematical text. For more information for mathematical annotation see `?plotmath`.

Add a legend.
```{r}
plot(Girth, pch='G',ylim=c(8,77), xlab='Tree number', ylab='', type='b', col='blue')
points(Volume, pch='V', type='b', col='red')
legend(x=2, y=70, legend=c('Girth', 'Volume'), pch=c('G','V'), col=c('blue','red'), bg='grey')
```


Adjusting Axes with `xlim` and `ylim`.

```{r}
plot(Girth, xlim=c(0,15), ylim=c(8,12))
```

Use `layout` for complicated plot layouts.

```{r}
A<-matrix(c(1,1,2,3,4,4,5,6), byrow=TRUE, ncol=2)
layout(A,heights=c(1/14,6/14,1/14,6/14))

oma.saved <- par("oma")
par(oma = rep.int(0, 4))
par(oma = oma.saved)
o.par <- par(mar = rep.int(0, 4))
for (i in seq_len(6)) {
	plot.new()
	box()
	text(0.5, 0.5, paste('Box no.',i), cex=3)
}
```


Always detach.

```{r}
detach(trees)
```






### The Power of the graphics device

Building a line graph from scratch.

```{r}
x = 1995:2005
y = c(81.1, 83.1, 84.3, 85.2, 85.4, 86.5, 88.3, 88.6, 90.8, 91.1, 91.3)
plot.new()
plot.window(xlim = range(x), ylim = range(y))
abline(h = -4:4, v = -4:4, col = "lightgrey")
lines(x, y, lwd = 2)
title(main = "A Line Graph Example",
		xlab = "Time",
		ylab = "Quality of R Graphics")
axis(1)
axis(2)
box()
```

Things to note:

- `plot.new` creates a new, empty, plotting device.
- `plot.window` determines the limits of the plotting region.
- `axis` adds the axes, and `box` the framing box. 
- The rest of the elements, you already know.

Rosette.

```{r}
n = 17
theta = seq(0, 2 * pi, length = n + 1)[1:n]
x = sin(theta)
y = cos(theta)
v1 = rep(1:n, n)
v2 = rep(1:n, rep(n, n))
plot.new()
plot.window(xlim = c(-1, 1), ylim = c(-1, 1), asp = 1)
segments(x[v1], y[v1], x[v2], y[v2])
box()
```



Arrows.

```{r}
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
arrows(.05, .075, .45, .9, code = 1)
arrows(.55, .9, .95, .075, code = 2)
arrows(.1, 0, .9, 0, code = 3)
text(.5, 1, "A", cex = 1.5)
text(0, 0, "B", cex = 1.5)
text(1, 0, "C", cex = 1.5)
```

Arrows as error bars.

```{r}
x = 1:10
y = runif(10) + rep(c(5, 6.5), c(5, 5))
yl = y - 0.25 - runif(10)/3
yu = y + 0.25 + runif(10)/3
plot.new()
plot.window(xlim = c(0.5, 10.5), ylim = range(yl, yu))
arrows(x, yl, x, yu, code = 3, angle = 90, length = .125)
points(x, y, pch = 19, cex = 1.5)
axis(1, at = 1:10, labels = LETTERS[1:10])
axis(2, las = 1)
box()
```


A histogram is nothing but a bunch of rectangle elements.

```{r}
plot.new()
plot.window(xlim = c(0, 5), ylim = c(0, 10))
rect(0:4, 0, 1:5, c(7, 8, 4, 3), col = "lightblue")
axis(1)
axis(2, las = 1)
```


Spiral Squares.

```{r}
plot.new()
plot.window(xlim = c(-1, 1), ylim = c(-1, 1), asp = 1)
x = c(-1, 1, 1, -1)
y = c( 1, 1, -1, -1)
polygon(x, y, col = "cornsilk")
vertex1 = c(1, 2, 3, 4)
vertex2 = c(2, 3, 4, 1)
for(i in 1:50) {
	x = 0.9 * x[vertex1] + 0.1 * x[vertex2]
	y = 0.9 * y[vertex1] + 0.1 * y[vertex2]
	polygon(x, y, col = "cornsilk")
}
```

Circles are just dense polygons.

```{r}
R = 1
xc = 0
yc = 0
n = 72
t = seq(0, 2 * pi, length = n)[1:(n-1)]
x = xc + R * cos(t)
y = yc + R * sin(t)
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
polygon(x, y, col = "lightblue", border = "navyblue")
```



Spiral- just a bunch of lines. 

```{r}
k = 5
n = k * 72
theta = seq(0, k * 2 * pi, length = n)
R = .98^(1:n - 1)
x = R * cos(theta)
y = R * sin(theta)
plot.new()
plot.window(xlim = range(x), ylim = range(y), asp = 1)
lines(x, y)
```



### Exporting a Plot

The pipile for exporting graphics is similar to the export of data. 
Instead of the `write.table` or `save` functinons, we will use the `pdf`, `tiff`, `png`, functions. Depending on the type of desired output.


Check and set the working directory.

```{r, eval=FALSE}
getwd()
setwd("/tmp/")
```

Export tiff.

```{r, eval=FALSE}
tiff(filename='graphicExample.tiff')
plot(rnorm(100))
dev.off()
```

Things to note:

- The `tiff` function tells R to open a .tiff file, and write the output of a plot.
- Only a single (the last) plot is saved.
- `dev.off` is close to close the tiff device, and return the plotting to the R console (or RStudio).

If you want to produce several plots, you can use a counter in the file's name.

```{r, evel=FALSE}
tiff(filename='graphicExample%d.tiff') #Creates a sequence of files
plot(rnorm(100))
boxplot(rnorm(100))
hist(rnorm(100))
dev.off()
```

To see the list of all open devices use `dev.list()`. 
To close __all__ device, (not the last one), use `graphics.off()`.


See `?pdf` and `?jpeg` for more info.




## The ggplot2 System

The philosophy of __ggplot2__ is very different from the __graphics__ device. 
Recall, in __ggplot2__, a plot is a object.
It can be queryied, it can be changed, and among other things, it can be plotted. 

The following is taken from [UCLA's idre](http://www.ats.ucla.edu/stat/r/seminars/ggplot2_intro/ggplot2_intro.htm).

A __ggplot2__ object will have the following elements:

- __Data__ are the variables mapped to aesthetic features of the graph.
- __Aes__ is the mapping between objects to their visualization.
- __Geoms__ are the objects/shapes you see on the graph.
- __Stats__ are stastical transformations that summarize data, such as the mean or confidence intervals.
- __Scales__ define which aesthetic values are mapped to data values. Legends and axes display these mappings.
- __Coordiante systems__ define the plane on which data are mapped on the graphic.
- __Faceting__ splits the data into subsets to create multiple variations of the same graph (paneling).

The `nlme::Milk` dataset has the protein level of various cows, at various times, with various diets.
```{r}
library(nlme)
data(Milk)
head(Milk)
```

```{r}
library(ggplot2)
ggplot(data = Milk, aes(x=Time, y=protein)) +
  geom_point()
```

Things to note:

- The `ggplot` function is the contructor of the __ggplot2__ object. If the object is not assigned, it is plotted.
- The `aes` argument tells R that the `Time` variable in the `Milk` data is the x axis, and protein is y.
- The `geom_point` defines the __Geom__, i.e., it tells R to plot the points as they are (and not lines, histograms, etc.).
- The __ggplot2__ object is build by compunding its various elements separated by the `+` operator.
- All the variables that we will need are assumed to be in the `Milk` data frame. This means that (a) the data needs to be a data frame (not a matrix for instance), and (b) we will not be able to use variables that are not in the `Milk` data frame.


Let's add some color. 

```{r}
ggplot(data = Milk, aes(x=Time, y=protein)) +
  geom_point(aes(color=Diet))
```

The `color` argument tells R to use the variable `Diet` as the coloring. If we wanted a fixed color, and not a variable dependent color, `color` would have been put __outside__ the `aes` function.

```{r}
ggplot(data = Milk, aes(x=Time, y=protein)) +
  geom_point(color="green")
```

Let's save the __ggplot2__ object so we can reuse it. Notice it is not plotted.

```{r}
p <- ggplot(data = Milk, aes(x=Time, y=protein)) +
  geom_point()
```

We can add __layers__ using the `+` operator. 
Here, we add a smoothing line.

```{r}
p + geom_smooth()
```

Things to note:

- The smoothing line is a layer added with the `geom_smooth()` function.
- Lacking any arguments, the new layer will inherit the `aes` of the original object, x and y variables in particular. 


To split the plot along some variable, we use faceting, done with the `facet_wrap` function.

```{r}
p + facet_wrap(~Diet)
```

We now add a layer of the mean of each `Diet` subgroup, connected by lines.

```{r}
p + stat_summary(aes(color=Diet), fun.y="mean", geom="line")
```

Things to note:

- `stat_summary` adds a statistical summary.
- The summary is applied along `Diet` subgroups, because of the `color=Diet` aesthetic.
- The summary to be applied is the mean, because of `fun.y="mean"`.
- The group means are connected by lines, because of the `geom="line"` argument.

What layers can be added using the __geoms__ family of functions?

- __geom_bar__: bars with bases on the x-axis.
- __geom_boxplot__: boxes-and-whiskers.
- __geom_errorbar__: T-shaped error bars.
- __geom_histogram__: histogram.
- __geom_line__: lines.
- __geom_point__: points (scatterplot).
- __geom_ribbon__: bands spanning y-values across a range of x-values.
- __geom_smooth__: smoothed conditional means (e.g. loess smooth).


To demonstrate the layers added with the __geoms__ functions, we start with a histogram.

```{r}
pro <- ggplot(Milk, aes(x=protein))
pro + geom_histogram()
```

A bar plot.

```{r}
ggplot(Milk, aes(x=Diet)) +
  geom_bar()
```

A scatter plot.

```{r}
tp <- ggplot(Milk, aes(x=Time, y=protein))
tp + geom_point()
```

A smooth regression plot, reusing the `tp` object.

```{r}
tp + geom_smooth()
```

And now, a simple line plot, reusing the `tp` object, and connecting lines along `Cow`.

```{r}
tp + geom_line(aes(group=Cow))
```

The line plot is completely incomprehensible. 
Better look at boxplots along time (even if ommiting the `Cow` information).
```{r}
tp + geom_boxplot(aes(group=Time))
```

We can do some statistics for each subgroup.
The following will compute the mean and standard errors (default of `stat_summary`) of `protein` at each time point.

```{r}
ggplot(Milk, aes(x=Time, y=protein)) +
  stat_summary()
```

Some popular statistical summaries, have gained their own functions:

- `mean_cl_boot`: mean and bootstapped confidence interval (default 95%).
- `mean_cl_normal`: mean and Guassian (t-distribution based) confidence interval (default 95%).
- `mean_dsl`: mean plus or minus standard deviation times some constant (default constant=2).
- `median_hilow`: median and outer quantiles (default outer quantiles = 0.025 and 0.975).

For less popular statistical summaries, we may specify the statistical function in `stat_summary`. The median is a first example.

```{r}
ggplot(Milk, aes(x=Time, y=protein)) +
  stat_summary(fun.y="median", geom="point")
```

We can also define our own statistical summaries.
```{r}
medianlog <- function(y) {median(log(y))}
ggplot(Milk, aes(x=Time, y=protein)) +
  stat_summary(fun.y="medianlog", geom="line")

```


__Scales__ define the actual mapping of an aesthetic's values to data values.

```{r}
ggplot(Milk, aes(x=protein, fill=Diet)) +
  geom_density(alpha=1/3) +
  scale_fill_hue()
```

Things to note:

- The `geom_density` function tells R to plot density plots.
- The `alpha=1/3` parameter controls the transparancy. Set to `alpha=1` for opaque, and `alpha=0` for transparent.
- The `scale_fill_hue` function, which is the default (thus can be omitted), tells R how to map factors to colors. 

Let's change the default color mapping. 

```{r}
ggplot(Milk, aes(x=protein, fill=Diet)) +
  geom_density(alpha=1/3) +
  scale_fill_hue(h.start=150)
```

The legend is controlled with the `guides` function.

```{r}
ggplot(Milk, aes(x=protein, fill=Diet)) +
  geom_density(alpha=1/3) +
  guides(fill="none")
```

__Faceting allows to split the plotting along some variable.
`face_wrap` tells R to compute the number of columns and rows of plots automatically. 

```{r}
ggplot(Milk, aes(x=protein, color=Diet)) +
  geom_density() +
  facet_wrap(~Time)
```

`facet_grid` forces the plot to appear alow rows or columns, using the `~` syntax.

```{r}
ggplot(Milk, aes(x=Time, y=protein)) +
  geom_point() +
  facet_grid(Diet~.)
```

To control the looks of the plot, __ggplot2__ uses __themes__.

```{r}
ggplot(Milk, aes(x=Time, y=protein)) +
  geom_point() +
  theme(panel.background=element_rect(fill="lightblue"))
```

```{r}
ggplot(Milk, aes(x=Time, y=protein)) +
  geom_point() +
  theme(panel.background=element_blank(),
        axis.title.x=element_blank())
```

Saving plots can be done using the `pdf` function, but possibly easier with the `ggsave` function.






## Interactive Graphics






## Shiny


## Bibliographic Notes
For the __graphics__ package, see @Rlanguage.
For __ggplot2__ see @ggplot2.

