---
output: html_document
editor_options: 
  chunk_output_type: console
---
# data.table {#datatable}

`data.table` is an excellent extension of the `data.frame` class.
If used as a `data.frame` it will look and feel like a data frame.
If, however, it is used with it's unique capabilities, it will prove faster and easier to manipulate. 
This is because `data.frame`s, like most of R objects, make a copy of themselves when modified. 
This is known as [passing by value](https://stackoverflow.com/questions/373419/whats-the-difference-between-passing-by-reference-vs-passing-by-value), and it is done to ensure that object are not corrupted if an operation fails (if your computer shuts down before the operation is completed, for instance). 
Making copies of large objects is clearly time and memory consuming.
A `data.table` can make changes in place. 
This is known as [passing by reference](https://stackoverflow.com/questions/373419/whats-the-difference-between-passing-by-reference-vs-passing-by-value), which is considerably faster than passing by value. 

Let's start with importing some freely available car sales data from [Kaggle](https://www.kaggle.com/orgesleka/used-cars-database).
```{r}
library(data.table)
library(magrittr)
auto <- fread('data/autos.csv')
```

```{r, eval=FALSE}
View(auto)
```

```{r}
dim(auto) #  Rows and columns
names(auto) # Variable names
class(auto) # Object class
file.info('data/autos.csv') # File info on disk
gdata::humanReadable(68439217)
object.size(auto) %>% print(units = 'auto') # File size in memory
```

Things to note:

- The import has been done with `fread` instead of `read.csv`. This is more efficient, and directly creates a `data.table` object.
- The import is very fast. 
- The data after import is slightly larger than when stored on disk (in this case). The extra data allows faster operation of this object, and the rule of thumb is to have 3 to 5 times more [RAM](https://en.wikipedia.org/wiki/Random-access_memory) than file size (e.g.: 4GB RAM for 1GB file)
- `auto` has two classes. It means that everything that expects a `data.frame` we can feed it a `data.table` and it will work.

Let's start with verifying that it behaves like a `data.frame` when expected.
```{r}
auto[,2] %>% head
auto[[2]] %>% head
auto[1,2] %>% head
```

But notice the difference between `data.frame` and `data.table` when subsetting multiple rows. Uhh!
```{r}
auto[1:3] %>% dim # data.table will exctract *rows*
as.data.frame(auto)[1:3] %>% dim # data.frame will exctract *columns*
```
Just use columns (`,`) and be explicit regarding the dimension you are extracting...


Now let's do some `data.table` specific operations.
The general syntax has the form `DT[i,j,by]`.
SQL users may think of `i` as `WHERE`, `j` as `SELECT`, and `by` as `GROUP BY`.
We don't need to name the arguments explicitly.
Also, the `Tab` key will typically help you to fill in column names.

```{r}
auto[,vehicleType,] %>% table # Exctract column and tabulate
auto[vehicleType=='coupe',,] %>% dim # Exctract rows 
auto[,gearbox:model,] %>% head # exctract column range
auto[,gearbox,] %>% table
auto[vehicleType=='coupe' & gearbox=='automatik',,] %>% dim # intersect conditions
auto[,table(vehicleType),] # uhh? why would this even work?!?
auto[, mean(price), by=vehicleType] # average price by car group
```


The `.N` operator is very useful if you need to count the length of the result. 
Notice where I use it:
```{r}
auto[.N,,] # will exctract the *last* row
auto[,.N] # will count rows
auto[,.N, vehicleType] # will count rows by type
```


You may concatenate results into a vector:
```{r}
auto[,c(mean(price), mean(powerPS)),]
```

This `c()` syntax no longer behaves well if splitting:
```{r}
auto[,c(mean(price), mean(powerPS)), by=vehicleType]
```

Use a `list()` instead of `c()`, within `data.table` commands:
```{r}
auto[,list(mean(price), mean(powerPS)), by=vehicleType]
```

You can add names to your new variables:
```{r}
auto[,list(Price=mean(price), Power=mean(powerPS)), by=vehicleType]
```

You can use `.()` to replace the longer `list()` command:
```{r}
auto[,.(Price=mean(price), Power=mean(powerPS)), by=vehicleType]
```

And split by multiple variables:
```{r}
auto[,.(Price=mean(price), Power=mean(powerPS)), by=.(vehicleType,fuelType)] %>% head
```

Compute with variables created on the fly:
```{r}
auto[,sum(price<1e4),] # Count prices lower than 10,000
auto[,mean(price<1e4),] # Proportion of prices lower than 10,000
auto[,.(Power=mean(powerPS)), by=.(PriceRange=price>1e4)] 
```

Things to note:

- The term `price<1e4` creates *on the fly* a binary vector of TRUE=1 / FALSE=0 for prices less than 10k and then sums/means this vector, hence `sum` is actually a count, and `mean` is proportion=count/total
- Summing all prices lower than 10k is done with the command `auto[price<1e4,sum(price),]`

You may sort along one or more columns
```{r}
auto[order(-price), price,] %>% head # Order along price. Descending
auto[order(price, -lastSeen), price,] %>% head# Order along price and last seen . Ascending and descending.
```


You may apply a function to ALL columns using a Subset of the Data using `.SD`
```{r}
count.uniques <- function(x) length(unique(x))
auto[,lapply(.SD, count.uniques), vehicleType]
```

Things to note:

- `.SD` is the data subset after splitting along the `by` argument. 
- Recall that `lapply` applies the same function to all elements of a list. In this example, to all columns of `.SD`.

If you want to apply a function only to a subset of columns, use the `.SDcols` argument
```{r}
auto[,lapply(.SD, count.uniques), by=vehicleType, .SDcols=price:gearbox]
```

## Make your own variables

It is very easy to compute new variables
```{r}
auto[,log(price/powerPS),] %>% head # This makes no sense
```

And if you want to store the result in a new variable, use the `:=` operator
```{r}
auto[,newVar:=log(price/powerPS),]
```

Or create multiple variables at once. 
The syntax `c("A","B"):=.(expression1,expression2)`is read "save the __list__ of results from expression1 and expression2 using the __vector__ of names A, and B".
```{r}
auto[,c('newVar','newVar2'):=.(log(price/powerPS),price^2/powerPS),]
```



## Join

__data.table__ can be used for joining.
A _join_ is the operation of aligning two (or more) data frames/tables along some index.
The index can be a single variable, or a combination thereof.

Here is a simple example of aligning age and gender from two different data tables:
```{r}
DT1 <- data.table(Names=c("Alice","Bob"), Age=c(29,31))
DT2 <- data.table(Names=c("Alice","Bob","Carl"), Gender=c("F","M","M"))
setkey(DT1, Names)
setkey(DT2, Names)
DT1[DT2,,] 
DT2[DT1,,] 
```

Things to note:

- A join with `data.tables` is performed by indexing one `data.table` with another. Which is the outer and which is the inner will affect the result.
- The indexing variable needs to be set using the `setkey` function.

There are several types of joins:

- __Inner join__: Returns the rows along the intersection of keys, i.e., rows that appear in __all__ data sets.
- __Outer join__: Returns the rows along the union of keys, i.e., rows that appear in __any__ of the data sets. 
- __Left join__: Returns the rows along the index of the "left" data set.
- __Right join__: Returns the rows along the index of the "right" data set.

Assuming `DT1` is the "left" data set, we see that `DT1[DT2,,]` is a right join, and `DT2[DT1,,]` is a left join.
For an inner join use the `nomath=0` argument:
```{r}
DT1[DT2,,,nomatch=0]
DT2[DT1,,,nomatch=0]
```



## Reshaping data

Data sets (i.e. frames or tables) may arrive in a "wide" form or a "long" form.
The difference is best illustrated with an example. 
The `ChickWeight` data encodes the weight of various chicks. It is "long" in that a variable encodes the time of measurement, making the data, well, simply long:
```{r}
ChickWeight %>%  head
```

The `mtcars` data encodes 11 characteristics of 32 types of automobiles. It is "wide" since the various characteristics are encoded in different variables, making the data, well, simply wide.

```{r}
mtcars %>% head
```

Most of _R_'s functions, with exceptions, will prefer data in the long format. 
There are thus various facilities to convert from one format to another. 
We will focus on the `melt` and `dcast` functions to convert from one format to another.

### Wide to long

`melt` will convert from wide to long. 

```{r}
dimnames(mtcars)
mtcars$type <- rownames(mtcars)
melt(mtcars, id.vars=c("type")) %>% head
```


Things to note:

- The car type was originally encoded in the rows' names, and not as a variable. We thus created an explicit variable with the cars' type using the `rownames` function.
- The `id.vars` of the `melt` function names the variables that will be used as identifiers. All other variables are assumed to be measurements. These can have been specified using their index instead of their name.
- If not all variables are measurements, we could have names measurement variables explicitly using the `measure.vars` argument of the `melt` function. These can have been specified using their index instead of their name.
- By default, the molten columns are automatically named `variable` and `value`.

We can replace the automatic namings using `variable.name` and `value.name`:
```{r}
melt(mtcars, id.vars=c("type"), variable.name="Charachteristic", value.name="Measurement") %>% head
```

### Long to wide

`dcast` will convert from long to wide:
```{r}
dcast(ChickWeight, Chick~Time, value.var="weight")
```

Things to note:

- `dcast` uses a formula interface (`~`) to specify the row identifier and the variables. The LHS is the row identifier, and the RHS for the variables to be created. 
- The measurement of each LHS at each RHS, is specified using the `value.var` argument.





## Bibliographic Notes
`data.table` has excellent online documentation. 
See  [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html).
See [here](https://rstudio-pubs-static.s3.amazonaws.com/52230_5ae0d25125b544caab32f75f0360e775.html) for __joining__. 
See [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html) for more on __reshaping__.
See [here](https://www.r-bloggers.com/intro-to-the-data-table-package/) for a comparison of the `data.frame` way, versus the `data.table` way.
For some advanced tips and tricks see [Andrew Brooks' blog](http://brooksandrew.github.io/simpleblog/articles/advanced-data-table/).

## Practice Yourself

1. Create a matrix of ones with `1e5` rows and `1e2` columns. Create a `data.table` using this matrix. 
    1. Replace the first column of each, with the sequence $1,2,3,\dots$. 
    1. Create a column which is the sum of all columns, and a $\mathcal{N}(0,1)$ random variable. 

2. Use the cars dataset used in this chapter from kaggle [Kaggle](https://www.kaggle.com/orgesleka/used-cars-database). 
    1. Import the data using the function `fread`. What is the class of your object? 
    1. Use `system.time()` to measure the time to sort along "seller". Do the same after converting the data to `data.frame`. Are data tables faster?


Also, see DataCamp's [Data Manipulation in R with data.table](https://www.datacamp.com/courses/data-manipulation-in-r-with-datatable), by Matt Dowle, the author of _data.table_ for more self practice.
