---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Multivariate Data Analysis {#multivariate}

The term "multivariate data analysis" is so broad and so overloaded, that we start by clarifying what is discussed and what is not discussed in this chapter.
Broadly speaking, we will discuss statistical _inference_, and leave more "exploratory flavored" matters like clustering, and visualization, to the Unsupervised Learning Chapter \@ref(unsupervised).

We start with an example.

```{example, label='icu'}
Consider the problem of a patient monitored in the intensive care unit. 
At every minute the monitor takes $p$ physiological measurements: blood pressure, body temperature, etc.
The total number of minutes in our data is $n$, so that in total, we have $n \times p$ measurements, arranged in a matrix. 
We also know the typical $p$-vector of typical measurements for this patient when healthy, denoted $\mu_0$.
```

Formally, let $y$ be single (random) measurement of a $p$-variate random vector.
Denote $\mu:=E[y]$.
Here is the set of problems we will discuss, in order of their statistical difficulty. 

- __Signal Detection__: 
a.k.a. _multivariate test_, or _global test_, or _omnibus test_.
Where we test whether $\mu$ differs than some $\mu_0$. 


- __Signal Counting__: 
a.k.a. _prevalence estimation_, or _$\pi_0$ estimation_. 
Where we count the number of entries in $\mu$ that differ from $\mu_0$.

- __Signal Identification__:
a.k.a. _selection_, or _multiple testing_.
Where we infer which of the entries in $\mu$ differ from $\mu_0$. 
In the ANOVA literature, this is known as a __post-hoc__ analysis, which follows an _omnibus test_.

- __Estimation__: 
Estimating the magnitudes of entries in $\mu$, and their departure from $\mu_0$. 
If estimation follows a _signal detection_ or _signal identification_ stage, this is known as _selective estimation_.



```{example, label='brain-imaging'}
Consider the problem of detecting regions of cognitive function in the brain using fMRI.
Each measurement is the activation level at each location in a brain's region.
If the region has a cognitive function, the mean activation differs than $\mu_0=0$ when the region is evoked.
```

```{example, label='genetics'}
Consider the problem of detecting cancer encoding regions in the genome.
Each measurement is the vector of the genetic configuration of an individual.
A cancer encoding region will have a different (multivariate) distribution between sick and healthy.
In particular, $\mu$ of sick will differ from $\mu$ of healthy.
```


```{example, label='regression'}
Consider the problem of the simplest multiple regression. 
The estimated coefficient, $\hat \beta$ are a random vector.
Regression theory tells us that its covariance is $Var[\hat \beta|X]=(X'X)^{-1}\sigma^2$, and null mean of $\beta$.
We thus see that inference on the vector of regression coefficients, is nothing more than a multivaraite inference problem.
```





## Signal Detection
Signal detection deals with the detection of the departure of $\mu$ from some $\mu_0$, and especially, $\mu_0=0$.
This problem can be thought of as the multivariate counterpart of the univariate hypothesis t-test. 

### Hotelling's T2 Test
The most fundamental approach to signal detection is a mere generalization of the t-test, known as _Hotelling's $T^2$ test_. 

Recall the univariate t-statistic of a data vector $x$ of length $n$:
\begin{align}
  t^2(x):= \frac{(\bar{x}-\mu_0)^2}{Var[\bar{x}]}= (\bar{x}-\mu_0)Var[\bar{x}]^{-1}(\bar{x}-\mu_0),
  (\#eq:t-test)
\end{align}
where $Var[\bar{x}]=S^2(x)/n$, and $S^2(x)$ is the unbiased variance estimator $S^2(x):=(n-1)^{-1}\sum (x_i-\bar x)^2$.

Generalizing Eq\@ref(eq:t-test) to the multivariate case: 
$\mu_0$ is a $p$-vector, $\bar x$ is a $p$-vector, and $Var[\bar x]$ is a $p \times p$ matrix of the covariance between the $p$ coordinated of $\bar x$. 
When operating with vectors, the squaring becomes a quadratic form, and the division becomes a matrix inverse. 
We thus have
\begin{align}
  T^2(x):= (\bar{x}-\mu_0)' Var[\bar{x}]^{-1} (\bar{x}-\mu_0),
  (\#eq:hotelling-test)
\end{align}
which is the definition of Hotelling's $T^2$ one-sample test statistic. 
We typically denote the covariance between coordinates in $x$ with $\hat \Sigma(x)$, so that 
$\widehat \Sigma_{k,l}:=\widehat {Cov}[x_k,x_l]=(n-1)^{-1} \sum (x_{k,i}-\bar x_k)(x_{l,i}-\bar x_l)$.
Using the $\Sigma$ notation, Eq.\@ref(eq:hotelling-test) becomes 
\begin{align}
  T^2(x):= n (\bar{x}-\mu_0)' \hat \Sigma(x)^{-1} (\bar{x}-\mu_0),
\end{align}
which is the standard notation of Hotelling's test statistic.

For inference, we need the null distribution of Hotelling's test statistic. For this we introduce some vocabulary^[This vocabulary is not standard in the literature, so when you read a text, you will need to verify yourself what the author means.]:

1. __Low Dimension__:
We call a problem _low dimensional_ if $n \gg p$, i.e. $p/n \approx 0$.
This means there are many observations per estimated parameter.
1. __High Dimension__: 
We call a problem _high dimensional_ if $p/n \to c$, where $c\in (0,1)$.
This means there are more observations than parameters, but not many.
1. __Very High Dimension__: 
We call a problem _very high dimensional_ if $p/n \to c$, where $1<c<\infty$.
This means there are less observations than parameters.

Hotelling's $T^2$ test can only be used in the low dimensional regime. 
For some intuition on this statement, think of taking $n=20$ measurements of $p=100$ physiological variables. 
We seemingly have $20$ observations, but there are $100$ unknown quantities in $\mu$. 
Say you decide that $\mu$ differs from $\mu_0$ based on the coordinate with maximal difference between your data and $\mu_0$. 
Do you know how much variability to expect of this maximum? 
Try comparing your intuition with a quick simulation.
Did the variability of the maximum surprise you? 
Hotelling's $T^2$ is not the same as the maximum, but the same intuition applies. 
This criticism is formalized in @bai1996effect. 
In modern applications, Hotelling's $T^2$ is rarely recommended.
Luckily, many modern alternatives are available. 
See @rosenblatt2016better for a review.


### Various Types of Signal to Detect
In the previous, we assumed that the signal is a departure of $\mu$ from some $\mu_0$. 
For vector-valued data $y$, that is distributed $\mathcal F$, we may define "signal" as any departure from some $\mathcal F_0$. 
This is the multivaraite counterpart of goodness-of-fit (GOF) tests. 

Even when restricting "signal" to departures of $\mu$ from $\mu_0$, "signal" may come in various forms:

1. __Dense Signal__: when the departure is in a large number of coordinates of $\mu$. 
1. __Sparse Signal__: when the departure is in a small number of coordinates of $\mu$.

Process control in a manufacturing plant, for instance, is consistent with a dense signal: if a manufacturing process has failed, we expect a change in many measurements (i.e. coordinates of $\mu$). 
Detection of activation in brain imaging is consistent with a dense signal: if a region encodes cognitive function, we expect a change in many brain locations (i.e. coordinates of $\mu$.)
Detection of disease encoding regions in the genome is consistent with a sparse signal: if susceptibility of disease is genetic, only a small subset of locations in the genome will encode it. 

Hotelling's $T^2$ statistic is best for dense signal. 
The next test, is a simple (and forgotten) test best with sparse signal.


### Simes' Test
Hotelling's $T^2$ statistic has currently two limitations: It is designed for dense signals, and it requires estimating the covariance, which is a very difficult problem.

An algorithm, that is sensitive to sparse signal and allows statistically valid detection under a wide range of covariances (even if we don't know the covariance) is known as _Simes' Test_.
The statistic is defined vie the following algorithm:

1. Compute $p$ variable-wise p-values: $p_1,\dots,p_j$.
1. Denote $p_{(1)},\dots,p_{(j)}$ the sorted p-values.
1. Simes' statistic is $p_{Simes}:=min_j\{p_{(j)} \times p/j\}$.
1. Reject the "no signal" null hypothesis at significance $\alpha$ if $p_{Simes}<\alpha$. 



### Signal Detection with R

We start with simulating some data with no signal.
We  will convince ourselves that Hotelling's and Simes' tests detect nothing, when nothing is present.
We will then generate new data, after injecting some signal, i.e., making $\mu$ depart from $\mu_0=0$. 
We then convince ourselves, that both Hotelling's and Simes' tests, are indeed capable of detecting signal, when present. 

Generating null data:

```{r}
library(mvtnorm)
n <- 100 # observations
p <- 18 # parameter dimension
mu <- rep(0,p) # no signal: mu=0
x <- rmvnorm(n = n, mean = mu)
dim(x)
lattice::levelplot(x) # just looking at white noise.
```

Now making our own Hotelling one-sample $T^2$ test using Eq.(\@ref(eq:hotelling-test)). 

```{r hotelling-function}
hotellingOneSample <- function(x, mu0=rep(0,ncol(x))){
  n <- nrow(x)
  p <- ncol(x)
  stopifnot(n > 5* p)
  bar.x <- colMeans(x)
  Sigma <- var(x)
  Sigma.inv <- solve(Sigma)
  T2 <- n * (bar.x-mu0) %*% Sigma.inv %*% (bar.x-mu0)
  p.value <- pchisq(q = T2, df = p, lower.tail = FALSE)
  return(list(statistic=T2, pvalue=p.value))
}
hotellingOneSample(x)
```

Things to note:

- `stopifnot(n > 5 * p)` is a little verification to check that the problem is indeed low dimensional. Otherwise, the $\chi^2$ approximation cannot be trusted.
- `solve` returns a matrix inverse. 
- `%*%` is the matrix product operator (see also `crossprod()`).
- A function may return only a single object, so we wrap the statistic and its p-value in a `list` object.

Just for verification, we compare our home made Hotelling's test, to the implementation in the __rrcov__ package. 
The statistic is clearly OK, but our $\chi^2$ approximation of the distribution leaves room to desire. 
Personally, I would never trust a Hotelling test if $n$ is not much greater than $p$, in which case the high-dimensional-statistics literature is worth consulting.

```{r}
rrcov::T2.test(x)
```

Now write our own Simes' test, and verify that it indeed does not find signal that is not there.
```{r}
Simes <- function(x){
  p.vals <- apply(x, 2, function(z) t.test(z)$p.value) # Compute variable-wise pvalues
  p <- ncol(x)
  p.Simes <- p * min(sort(p.vals)/seq_along(p.vals)) # Compute the Simes statistic
  return(c(pvalue=p.Simes))
}
Simes(x)
```

And now we verify that both tests can indeed detect signal when present. Are p-values small enough to reject the "no signal" null hypothesis?
```{r}
mu <- rep(x = 10/p,times=p) # inject signal
x <- rmvnorm(n = n, mean = mu)
hotellingOneSample(x)
Simes(x)
```

... yes. All p-values are very small, so that all statistics can detect the non-null distribution.




## Signal Counting 
There are many ways to approach the _signal counting_ problem. 
For the purposes of this book, however, we will not discuss them directly, and solve the signal counting problem via the solution to a signal identification problem. 
The rational is the following: if we know __where__ $\mu$ departs from $\mu_0$, we only need to count coordinates to solve the signal counting problem. 

We now make the previous argument a little more accurate.
Assume you have a selection/identification algorithm, that selects coordinates in $\mu$.
Denote with $R(\alpha)$ the number of selected coordinates, where $\alpha$ is the coordinate-wise false positive rate. 
Then $R(\alpha)$ includes approximately $\alpha p$ false positives. 
Denote by $p_0$ the number of coordinates that do not carry signal.
Then $p_0 \approx p-(R(\alpha)-p_0 \, \alpha)$.
Equating the two we have $$\hat p_0=\frac{p-R(\alpha)}{1-\alpha}.$$
The number of coordinates in $\mu$ that truly carry signal is thus approximately $p-\hat p_0$, which is a solution to the signal counting problem. 


## Signal Identification {#identification}

The problem of _signal identification_ is also known as a _selection problem_, or more commonly as _multiple testing_ problem.
In the ANOVA literature, an identification stage will typically follow a detection stage.
These are known as the _omnibus F test_, and _post-hoc_ tests, respectively.
In the multiple testing literature it is quite rare to see a preliminary detection stage, because it is assumed that signal is there; it is merely a problem of how much, and where?
I.e., counting and selecting, respectively.

The first question when approaching a multiple testing problem is "what is an error"?
Is an error declaring a coordinate in $\mu$ to be different than $\mu_0$ when it is actually not?
Is an error an overly high proportion of falsely identified coordinates?
The former is known as the _family wise error rate_ (FWER), and the latter as the _false discovery rate_ (FDR).

```{remark}
These types of errors have many names in many communities.
See the Wikipedia entry on [ROC](https://en.wikipedia.org/wiki/Receiver_operating_characteristic) for a table of the (endless) possible error measures. 
```



### Signal Identification in R

One (of many) ways to do signal identification with statistical guarantees involves the `stats::p.adjust` function.
The function takes as inputs a $p$-vector of the variable-wise __p-values__. 
Why do we start with variable-wise p-values, and not the full data set?

a. Because we want to make inference variable-wise, so it is natural to start with variable-wise statistics. 
a. Because we want to avoid dealing with covariances if possible. Computing variable-wise p-values does not require estimating covariances. 

We start be generating some high-dimensional multivariate data and computing the coordinate-wise (i.e. hypothesis-wise) p-value.

```{r}
library(mvtnorm)
n <- 1e1
p <- 1e2
mu <- rep(0,p)
x <- rmvnorm(n = n, mean = mu)
dim(x)
lattice::levelplot(x)
```

We now compute the pvalues of each coordinate.
We use a coordinate-wise t-test. 
Why a t-test? Because for the purpose of demonstration we want a simple test. In reality, you may use any test that returns valid p-values. 

```{r}
t.pval <- function(y) t.test(y)$p.value
p.values <- apply(X = x, MARGIN = 2, FUN = t.pval) 
plot(p.values, type='h')
```

Things to note:

- `t.pval` is a function that merely returns the p-value of a t.test.
- We used the `apply` function to apply the same function to each column of `x`. 
- `MARGIN=2` tells `apply` to compute over columns and not rows. 
- The output, `p.values`, is a vector of 100 p-values. 

We are now ready to do the identification, i.e., find which coordinate of $\mu$ is different than $\mu_0=0$.
The workflow for identification has the same structure, regardless of the desired error guarantees: 

1. Compute an `adjusted p-value`. 
1. Compare the adjusted p-value to the desired error level.

If we want $FWER \leq 0.05$, meaning that we allow a $5\%$ probability of making any mistake, we will use the `method="holm"` argument of `p.adjust`. 
Recalling that our data currently has no signal we verify that, indeed, nothing is selected. 

```{r holm}
alpha <- 0.05
p.values.holm <- p.adjust(p.values, method = 'holm' )
which(p.values.holm < alpha)
```

If we want $FDR \leq 0.05$, meaning that we allow the proportion of false discoveries to be no larger than $5\%$, we use the `method="BH"` argument of `p.adjust`.
Again, nothing is selected because there is nothing to select.

```{r bh}
alpha <- 0.05
p.values.BH <- p.adjust(p.values, method = 'BH' )
which(p.values.BH < alpha)
```

We now inject some strong signal in $\mu$ just to see that the process works.
We will artificially inject signal in the first 10 coordinates.

```{r}
mu[1:10] <- 2 # inject signal in first 10 variables
x <- rmvnorm(n = n, mean = mu) # generate data
p.values <- apply(X = x, MARGIN = 2, FUN = t.pval) 
p.values.BH <- p.adjust(p.values, method = 'BH' )
which(p.values.BH < alpha)
```

Indeed- we are now able to detect that the first coordinates carry signal, because their respective coordinate-wise null hypotheses have been rejected. 


## Signal Estimation (*)

The estimation of the elements of $\mu$ is a seemingly straightforward task.
This is not the case, however, if we estimate only the elements that were selected because they were significant (or any other data-dependent criterion).
Clearly, estimating only significant entries will introduce a bias in the estimation. 
In the statistical literature, this is known as _selection bias_.
Selection bias also occurs when you perform inference on regression coefficients after some model selection, say, with a lasso, or a forward search^[You might find this shocking, but it does mean that you cannot trust the `summary` table of a model that was selected from a multitude of models.].

Selective inference is a complicated and active research topic so we will not offer any off-the-shelf solution to the matter.
The curious reader is invited to read @rosenblatt2014selective, @javanmard2014confidence, or [Will Fithian's](http://www.stat.berkeley.edu/~wfithian/) PhD thesis [@fithian2015topics] for more on the topic. 









## Bibliographic Notes
For a general introduction to multivariate data analysis see @anderson2004introduction.
For an R oriented introduction, see @everitt2011introduction.
For more on the difficulties with high dimensional problems, see @bai1996effect.
For some cutting edge solutions for testing in high-dimension, see @rosenblatt2016better and references therein.
Simes' test is not very well known. It is introduced in @simes1986improved, and proven to control the type I error of detection under a PRDS type of dependence in @benjamini2001control.
For more on multiple testing, and signal identification, see @goeman2014multiple.
For more on the choice of your error rate see @rosenblatt2013practitioner.



## Practice Yourself

1. Generate multivariate data with:
    ```{r}
set.seed(3)
mu<-rexp(50,6)
multi<-  rmvnorm(n = 100, mean = mu) 
```
    a. Use Hotelling's test to determine if $\mu$ equals $\mu_0=0$. Can you detect the signal? 
    a. Perform t.test on each variable and extract the p-value. Try to identify visually the variables which depart from $\mu_0$.
    a. Use `p.adjust` to identify in which variables there are any departures from $\mu_0=0$. Allow 5% probability of making any false identification.
    a. Use `p.adjust` to identify in which variables there are any departures from $\mu_0=0$. Allow a 5% proportion of errors within identifications.

1. Generate multivariate data from two groups: `rmvnorm(n = 100, mean = rep(0,10))` for the first, and `rmvnorm(n = 100, mean = rep(0.1,10))` for the second.
    a. Do we agree the groups differ?
    b. Implement the two-group Hotelling test described in Wikipedia: (https://en.wikipedia.org/wiki/Hotelling%27s_T-squared_distribution#Two-sample_statistic).
    c. Verify that you are able to detect that the groups differ. 
    d. Perform a two-group t-test on each coordinate. On which coordinates can you detect signal while controlling the FWER? On which while controlling the FDR? Use `p.adjust`.

1. Return to the previous problem, but set `n=9`. Verify that you cannot compute your Hotelling statistic. 
